% Acquisition, Reconstruction, and World Models for Physical AI Data Generation
% Position / vision paper draft (v0.1)
% Target venue: HotOS-style position track, HPG, MLSys, or arXiv preprint.

\documentclass[10pt,journal,onecolumn]{IEEEtran}

\usepackage{graphicx}
\usepackage{amsmath,amssymb}
\usepackage{array}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{cite}
\usepackage{xcolor}
\usepackage{url}
\usepackage{tikz}
\usepackage{pgfplots}
\usetikzlibrary{positioning,arrows.meta,shapes.geometric,backgrounds,fit,calc}
\usepgfplotslibrary{fillbetween}
\pgfplotsset{compat=1.18}

% Academic paragraph convention: keep normal paragraph indentation, but do not
% indent the first paragraph after section-like headings.
\makeatletter
\let\vkgsplat@afterheading\@afterheading
\def\@afterheading{%
  \@afterindentfalse
  \vkgsplat@afterheading}
\makeatother

\newcommand{\todo}[1]{\textcolor{red}{[TODO: #1]}}
\newcommand{\note}[1]{\textcolor{blue}{[NOTE: #1]}}
\definecolor{plotblue}{RGB}{30,144,255}
\definecolor{plotcyan}{RGB}{0,180,190}
\definecolor{plotgreen}{RGB}{0,160,90}
\definecolor{plotyellow}{RGB}{235,180,35}
\definecolor{plotred}{RGB}{210,70,70}

\title{Acquisition, Reconstruction, and World Models:\\A Compute-First Architecture for Physical AI Data Generation}

\author{%
    Chien-Ping Lu\\[0.35em]
    \normalsize May 4, 2026%
}

\begin{document}
\maketitle

\begin{abstract}
\setlength{\parindent}{0pt}
\noindent
Physical AI needs enormous amounts of data, but the field is caught between two
imperfect stories. Classical simulation promises physical correctness but is
expensive, graphics-hardware-bound, and difficult to scale across AI
accelerator fleets. Generative world models promise cheap interactive futures,
but their predictions are only physical inside a validated envelope. This paper
argues for a third architecture: \emph{physical acquisition plus explicit state
plus scalable reconstruction and prediction}. Calibrated simulators, sensors,
and low-sample path tracers provide sparse physical evidence; persistent 3D/4D
state stores geometry, radiance, motion, visibility, variance, and confidence;
classical filters, Gaussian splatting, and neural/world-model components then
denoise, supersample, interpolate, predict, and propose counterfactuals. The
system returns to a physical or sensor oracle when uncertainty, rare events, or
task risk exceed the learned model's validation envelope. Crucially, synthetic
data generation and model training form a closed loop: the current robot model
acts in the simulated or reconstructed world, its failures and uncertainty drive
new acquisitions, and the resulting synthetic data revises the next model.
This framing explains several otherwise separate trends. DLSS-class rendering
already treats rasterization and ray tracing as sparse acquisition followed by
tensor-side reconstruction~\cite{nvidia-dlss,liu2020nsrr,nvidia-dlss3,nvidia-dlss35rr}.
However, DLSS is compute-heavy in implementation but proprietary and
graphics-stack-packaged in deployment. vkGSplat therefore treats integrated
3DGS/radiance state as the open DLSS-class replacement layer for SDG: a
device-resident reconstruction memory that consumes sparse seeds and serves
denoising, temporal reuse, super-resolution, frame interpolation, and
confidence-gated oracle refresh.
Robotics SDG is throughput-bound and latency-tolerant, so it can trade
fixed-function graphics latency for compute portability. More importantly, SDG
has an amortized world-building phase that games do not: it can spend seconds,
minutes, or hours establishing static geometry, materials, lighting priors,
BVHs, Gaussian maps, and high-bounce radiance caches before running many
simulation and acquisition episodes. World-model systems
such as Genie and Cosmos are compatible with, rather than replacements for,
calibrated physics and sensors when they act as priors, rollout generators, or
proposal mechanisms~\cite{ha2018,bruce2024,deepmind2024genie2,nvidia2024cosmos}.
We use \textbf{vkGSplat} as the reference implementation of this philosophy. The
name is intentional: \textbf{vk} means the system preserves Vulkan as the
application contract while supporting a CUDA software execution path, and
\textbf{GSplat} means 3D Gaussian Splatting becomes the open reconstruction
layer that replaces the architectural role of proprietary DLSS for SDG. The
architecture is acquisition-backend neutral: a conventional Vulkan renderer on
graphics hardware is a valid and often best acquisition oracle. vkGSplat adds a
compute-first Vulkan path that preserves the API as a portable application
contract while lowering visual acquisition and reconstruction to CUDA-class
kernels. The CUDA software path is useful when data generation, reconstruction,
training, and world-model inference should remain on tensor-rich accelerators,
or when the renderer must expose custom sample statistics and explicit Gaussian
state updates. vkGSplat is not the whole thesis; it is the concrete test of
whether physical-evidence acquisition, explicit Gaussian state, PyTorch-centered
orchestration, and tensor-rich hardware can form a practical data-generation
substrate. As a falsifiable milestone rather than a reported result, we target
\textbf{$\geq 100$ images\slash s\slash H100} at $1024{\times}1024$ from a
captured 3DGS scene, \textbf{$\leq 0.5$ J\slash image} on H100-class hardware,
and \textbf{within $2{\times}$ of an RTX-class graphics-hardware baseline (L40)
on dollars per image}, while preserving downstream sim-to-real task accuracy.
\end{abstract}

\begin{IEEEkeywords}
Physical AI, synthetic data generation, world models, acquisition and
reconstruction, DLSS-class reconstruction, Vulkan, software rendering, 3D
Gaussian splatting, robotics, AI accelerators.
\end{IEEEkeywords}

% =====================================================================
\section{Introduction}
% =====================================================================

Physical AI data generation is often framed as a choice between two extremes.
One extreme is calibrated simulation: physics engines, path tracers, sensor
models, and digital twins that try to measure the virtual world with known error
bars. The other extreme is learned world modeling: generative systems that
predict plausible futures, transform scenes, and create interactive rollouts
without explicitly replaying every physical process. Both are necessary, and
neither is sufficient alone. Physics without reconstruction is too expensive to
scale; generation without physical acquisition can drift away from the regimes
where robotics needs trust.

This paper takes a different position. The right unit of design is not a
renderer, a simulator, or a world model by itself. It is an
\emph{acquisition--state--reconstruction loop}. A calibrated simulator,
physical sensor, rasterizer, or low-sample path tracer acquires sparse evidence
about the world. An explicit state representation stores that evidence across
space, time, views, modalities, uncertainty, and task context. Reconstruction
and world-model modules then denoise, supersample, interpolate, fill missing
views, predict futures, and propose counterfactuals. When the state is
under-supported --- low sample count, shallow path-depth evidence, high
variance, disocclusion, distribution shift, or safety-critical task risk --- the
system returns to a physical or sensor oracle. Put simply: AI should reduce the
cost of acquiring, storing, reconstructing, and querying physical truth; it
should not silently replace physical truth.

This is not an open-loop dataset factory. The robot or embodied agent being
trained is part of the data-generation loop. A current perception, policy, or
world model acts in the simulated or reconstructed world; its rollouts expose
uncertain states, failures, near misses, and rare contacts; vkGSplat then
acquires or reconstructs the evidence needed to turn those events into new
training examples. The revised model changes the next rollout distribution.
Synthetic data generation and model training are therefore coupled, not staged as
``render once, train later.''

This loop also changes the time structure of rendering. A game renderer must
rebuild enough of the world to satisfy the next display frame. A SDG system can
first build a world model: static geometry, materials, light-transport priors,
BVHs, 3DGS state, and sensor calibration can be computed with generous budgets
before any policy rollout or synthetic camera trajectory begins. The online
simulation loop then updates dynamic state --- robot motion, movable objects,
contact, deformables, light changes, material edits, and occlusions --- against
that amortized static base. This is a qualitative difference, not merely a
slower frame budget. It means vkGSplat should optimize the whole lifecycle,
not just per-frame rendering.

Modern graphics already points in this direction. In DLSS-class real-time
pipelines, rasterization and ray tracing increasingly produce only a
low-resolution, temporally jittered, spatially sparse seed. Tensor-side
reconstruction performs super-resolution, anti-aliasing, denoising, and frame
generation~\cite{nvidia-dlss,liu2020nsrr,nvidia-dlss3}. Ray Reconstruction makes
the same point for path-traced lighting: sparse physical rays acquire evidence,
then a neural model reconstructs the missing samples~\cite{nvidia-dlss35rr}.
The important lesson for Physical AI is not that a denoiser is magically
physical. It is that low-sample acquisition can remain grounded in the correct
transport process while reconstruction amortizes the cost. Lowering samples per
pixel preserves the target integral in expectation; lowering bounce depth
changes the physical model and must be tracked as bias.

This distinction is central to vkGSplat. DLSS is fundamentally a compute
reconstruction pipeline --- it runs on tensor hardware and consumes motion,
depth, exposure, and history --- but it is packaged as a proprietary graphics
feature inside the RTX/game-rendering stack. That packaging is rational for
games, yet it is the wrong boundary for AIDC-native synthetic data generation.
vkGSplat keeps the useful architectural lesson and changes the substrate:
integrated 3DGS/radiance state replaces the role of DLSS as the reconstruction
layer, while remaining open, scene-aware, persistent across views, and directly
connected to CUDA/PyTorch training workflows.

World-model research fits the same loop when its role is stated carefully.
Classical world models learn compact predictive dynamics for control
\cite{ha2018}; Genie-style and Cosmos-style systems scale toward interactive or
generative environments for embodied agents
\cite{bruce2024,deepmind2024genie2,nvidia2024cosmos}. These models can replace
many expensive simulator calls inside a validation envelope: rollouts,
variations, priors, proposals, and counterfactuals. They should not replace the
calibrated physics, sensor, or rendering oracle when the output is a measurement
whose errors matter. A useful world model proposes; the acquisition oracle
checks, refreshes, or corrects when confidence is weak.

The hardware implication is just as important. A generic architecture should
allow visual acquisition to come from the best available source: a physical
camera, an offline path tracer, a conventional Vulkan renderer running on
graphics hardware, or a software renderer running on AI compute. Today's leading
SDG stacks --- NVIDIA Omniverse\slash Isaac~Sim\slash Replicator and Cosmos
\cite{nvidia2024cosmos}, Unreal-based platforms, Blender-based pipelines
\cite{denninger2019blenderproc,greff2022kubric} --- usually use the hardware
Vulkan or RTX path, and that is the right baseline when those GPUs are available
and co-location is not a concern. The open question is whether the same
acquisition contract can also execute efficiently on tensor-rich AI accelerators
(NVIDIA~H100\slash B200, Google~TPU, AWS Trainium, and related chips), which
dominate FLOPS, memory bandwidth, model deployment, and procurement. If
acquisition is a sparse seed and reconstruction is the scalable value center,
then a compute-first implementation becomes plausible. The question is not
whether graphics hardware disappears, but when the Vulkan execution contract
should be allowed to move off fixed-function graphics units.

There is also a blunt economic motivation. The AI data center (AIDC) has become
a hundred-billion-dollar capital program: Microsoft described an approximately
\$80B FY2025 investment in AI-enabled datacenters, Alphabet guided to roughly
\$75B of 2025 capital expenditures, and Meta raised its 2025 capital-expenditure
outlook to \$70--72B while warning that 2026 growth would be driven primarily by
infrastructure costs~\cite{microsoft-ai-datacenter-2025,alphabet-q4-2024-capex,meta-q3-2025-capex}.
Industry analysts describe the same pressure at the data-center level: Omdia's
Data Center World 2026 material frames the period as a ``trillion dollar''
AIDC boom, estimates \$6.07T of total IT spending in 2026, reports that
72.9\% of surveyed organizations expect IT-budget growth in 2026, and projects
data-center investment approaching \$1.9T by 2030 under a 20\% CAGR scenario
\cite{omdia-dcw-oas-2026}. The same deck points to rack-scale AI systems,
800VDC power distribution, BESS, self-generation, and liquid cooling as
near-term data-center design pressures. We use this as market and infrastructure
evidence, not as a law of nature. The implication is that SDG architecture must
fit a capital program that is already large, power-constrained, and
operations-heavy.
In that world, the goal is not to beat NVIDIA's RTX approach on its own terms.
RTX/Omniverse/Isaac-style SDG is the natural baseline when a graphics fleet is
already the production substrate. The problem is that many organizations are
already locked into a massive AIDC buildout, and disrupting that buildout to add
a parallel graphics island is economically and operationally expensive. The
economic bet behind vkGSplat is therefore simpler: make synthetic data
generation a normal AI data-center workload, scheduled, containerized,
monitored, billed, and optimized on the same accelerator fleet that already runs
training and inference.

This also reframes the comparison with NVIDIA's ecosystem. NVIDIA has a
powerful heterogeneous answer: Omniverse and OVX systems for physically
realistic simulation, graphics, and industrial digital twins, and DGX/DGX Cloud
for AI-factory training and inference~\cite{nvidia-ovx,nvidia-dgx-cloud}. That
split is technically coherent, and it is the correct reference architecture for
sites that can justify a dedicated graphics/simulation cluster beside the AI
cluster. vkGSplat argues for the other branch of the decision tree: a unified
software-defined SDG stack that reuses the massive compute resources already
being installed for AI data centers. The point is not that Omniverse, OVX, or
RTX are weak. The point is that two specialized clusters create a second
economic surface: separate capital planning, utilization targets, power and
cooling envelopes, driver stacks, job schedulers, telemetry, and data movement
between simulation and training. A compute-first Vulkan path is attractive
precisely because it tries to collapse that split into one programmable AI
infrastructure substrate.

Concretely, making SDG AIDC-native requires addressing three NVIDIA-shaped
frictions, not merely replacing one renderer. First, OVX/DGX segregation turns
simulation/graphics and training/inference into different production substrates.
Second, fixed-function graphics hardware ties acquisition to RTX-style
availability, driver queues, and image-oriented contracts even when the
downstream consumer is a tensor pipeline. Third, DLSS-class reconstruction
shows where much of the value has moved, but it is proprietary and
product-tuned for games rather than an open SDG evidence contract. vkGSplat
treats these as boundary problems. CUDA software rendering is the response to
fixed-function graphics dependence: it preserves the Vulkan acquisition
contract while allowing the work to run on AI accelerators without graphics
blocks. Integrated 3DGS reconstruction is the response to the DLSS boundary:
instead of a closed image-space upscaler, vkGSplat exposes a persistent
Gaussian/radiance state for denoising, temporal reuse, super-resolution, frame
interpolation, and optional learned residuals. Vulkan, RTX, Omniverse, and
OVX/DGX paths remain compatible where they are the right substrate, but the same
seed-buffer and explicit-state reconstruction stack can also live inside
CUDA/PyTorch on the AI cluster.

\textbf{vkGSplat} is the concrete reference implementation for this broader
architecture. It preserves Vulkan as an application boundary, treats hardware
Vulkan output as a first-class acquisition source, and adds a CUDA software path
for seed generation and reconstruction when the pipeline benefits from AI-compute
co-location. It also explores persistent 3D/4D Gaussian state as an explicit
memory for physical evidence. The name carries the system boundary: \emph{vk}
denotes Vulkan compatibility through a CUDA software path, while \emph{GSplat}
denotes 3D Gaussian Splatting as the open DLSS-class integration layer for
denoising, temporal reuse, super-resolution, frame interpolation, and radiance
state. The paper therefore has two layers. The first is the philosophy:
Physical AI data generation should combine calibrated acquisition, explicit
state, learned reconstruction, and uncertainty-gated oracle refresh. The second
is the system: vkGSplat tests whether this philosophy can be implemented as a
compute-first Vulkan path on AI accelerators.

\textbf{Relation to the Gaussian reconstruction note.} This paper defines the
world-modeling and compute-first SDG system boundary: physical acquisition,
backend-neutral visual acquisition, hardware Vulkan ingestion, software CUDA seed
buffers, explicit state, world-model compatibility, backend portability, and the
hardware-allocation thesis. The companion technical
note on integrated 3DGS reconstruction owns the detailed question of whether one
persistent 3D/4D Gaussian state can unify Monte Carlo denoising, radiance
caching, temporal AA, super-resolution, and frame interpolation. In other words,
this paper treats 3DGS as an explicit-state bridge and possible reconstruction
representation; the companion note works out the Gaussian-state math and CUDA
experiments.

\textbf{Contributions.} This is a position paper about Physical AI data
generation, with \textbf{vkGSplat} as the reference system. We make ten claims:

\begin{enumerate}
    \item \textbf{World-modeling philosophy.} Physical AI data generation
    should be organized as physical acquisition, explicit state, scalable
    reconstruction or prediction, and uncertainty-gated oracle refresh. Learned
    world models can replace simulator calls inside a validation envelope, but
    calibrated physics and sensor models remain the measurement authority
    (Sections~\ref{sec:physical-sim-sensors} and~\ref{sec:pytorch-world-hub}).

    \item \textbf{Graphics model.} Modern graphics is no longer best
    understood as a monolithic raster pipeline. It is an acquisition
    front-end --- primary visibility, shadow or light visibility,
    secondary visibility, depth, motion, and sparse radiance --- followed
    by reconstruction. Much of the programmable work already behaves like
    screen-space compute over G-buffers and history buffers; graphics shaders
    are pipeline-scheduled kernels, while compute shaders expose explicit
    load/store, workgroup, and user-defined data-layout control. vkGSplat moves
    one step further by lifting reconstruction out of screen space into
    persistent 3D/4D Gaussian, radiance-cache, and world-state address spaces
    (Section~\ref{sec:acqrec}).

    \item \textbf{3DGS as the open DLSS-class integration layer.}
    DLSS is compute reconstruction packaged as a proprietary graphics-stack
    feature. vkGSplat replaces that architectural role, not by cloning DLSS,
    but by using integrated 3DGS/radiance state as an open, persistent,
    scene-aware reconstruction layer for denoising, TAA, super-resolution,
    frame interpolation, radiance-cache reuse, and confidence-gated oracle
    refresh (Section~\ref{sec:dlss}).

    \item \textbf{Allocation law.} The modernized Amdahl argument uses the
    allocation objective
    $T(R,s,x)=(1-s)/(1+(R-1)x)+s/(1-x)$, where $x$ is the budget assigned to
    specialized graphics logic, $R$ is its bounded-stage efficiency advantage,
    and $s$ is the value-scalable share. It implies that fixed-function
    specialization is justified only when $R>R_c(s)=1/(1-s)$. As DLSS-like
    reconstruction, integrated 3DGS state, world models, and training push $s$
    toward one, the required fixed-function advantage diverges. Bounded
    visibility becomes acquisition, while reconstruction, prediction, and
    world-model inference remain scalable stages that should stay on
    programmable compute (Section~\ref{sec:amdahl-main} and
    Appendix~\ref{app:amdahl}).

    \item \textbf{Workload.} Robotics SDG is throughput-bound and
    latency-tolerant. A renderer that produces images in tens to hundreds
    of milliseconds is acceptable if it sustains high aggregate throughput
    and high quality. This relaxes the constraint that historically
    forced the use of fixed-function graphics hardware
    (Section~\ref{sec:workload}).

    \item \textbf{Economic packaging.} The goal is not primarily to beat
    RTX-class SDG on per-ray or per-frame graphics efficiency. It is to package
    SDG as a normal AI data-center workload whenever the organization already
    operates large AI accelerator fleets and cannot economically disrupt that
    buildout with a separate graphics island. The relevant comparison is fleet
    procurement, utilization, software qualification, data movement, cluster
    scheduling, and organizational fit (Section~\ref{sec:economic-case}).

    \item \textbf{World-building lifecycle.} SDG differs from games not only
    quantitatively but qualitatively: an expensive static world-building phase
    can be amortized over many later simulation and acquisition episodes.
    Geometry, materials, acceleration structures, high-bounce light transport,
    Gaussian maps, and radiance caches can be built before simulation begins,
    while dynamic lights, objects, and materials become explicit invalidation
    and refresh events rather than per-frame renderer obligations
    (Section~\ref{sec:world-building-lifecycle}).

    \item \textbf{Reference implementation.} vkGSplat preserves Vulkan as the
    application contract, accepts hardware Vulkan as a valid acquisition path,
    and adds a compute-first software acquisition and reconstruction backend.
    Its name summarizes the bet: \emph{vk} is Vulkan preserved through CUDA
    software execution, and \emph{GSplat} is 3DGS replacing DLSS's architectural
    role as an open reconstruction layer. The point is not merely software
    rendering, but an explicit bridge between physical evidence, Gaussian
    state, and tensor-side reconstruction
    (Section~\ref{sec:design}).

    \item \textbf{Ray-tracing seed path before 3DGS.}
    The active v1 implementation targets low-sample, Vulkan-ray-tracing
    shaped seed frames before returning to 3DGS. This keeps the project
    focused on taking existing Vulkan programs, preserving SPIR-V and
    camera/depth/primitive contracts, and lowering reconstruction work
    to CUDA-friendly tile kernels. The earlier tile-based 3DGS renderer
    remains available as an optional backend, but no longer gates the
    main milestones (Section~\ref{sec:v1}).

    \item \textbf{Hardware-agnostic backend.} The compute IR underlying
    the renderer can be retargeted from CUDA to SYCL, Triton, or
    AI-accelerator native APIs, broadening deployment to any chip with
    sufficient compute and memory bandwidth (Section~\ref{sec:backends}).
\end{enumerate}

\textbf{Reference implementation goals.} The implementation goal is not merely to run a
toy renderer on CUDA; it is to make existing Vulkan programs and
compute backends coexist behind one stable contract. Concretely, vkGSplat
aims to:

\begin{enumerate}
    \item \textbf{Preserve Vulkan as the application boundary.}
    Applications should continue to create Vulkan pipelines, descriptor
    sets, command buffers, SPIR-V modules, acceleration structures, and
    frame resources. The compatibility burden belongs inside vkGSplat,
    not in per-application ports.

    \item \textbf{Make SPIR-V and backend kernels share one ABI.}
    Every captured shader module should lower to an explicit kernel
    interface describing entry points, descriptor sets, bindings,
    storage classes, push constants, built-ins, and opaque resources
    such as TLAS handles. CPU reference kernels, CUDA kernels, and
    portability backends must consume the same interface rather than
    maintaining parallel hand-written bindings.

    \item \textbf{Use CUDA as the production backend.} The long-term
    target remains NVIDIA data-center CUDA execution, including
    CUDA-friendly tile scheduling, software BVH traversal, temporal
    reconstruction, denoising, and eventually direct Vulkan--CUDA
    synchronization and memory interop.

    \item \textbf{Use CPU and Metal as correctness backends.} CPU
    kernels define the portable reference behavior. Metal on Apple
    Silicon is a local native-GPU check that the contracts are not
    accidentally CUDA-only before the CUDA implementation is complete.

    \item \textbf{Keep the renderer tile based.} Rasterization,
    ray-tracing seed generation, reprojection, denoising, and optional
    3DGS blending should all be organized as tile- or block-local work
    so the algorithms map naturally to CUDA thread blocks and can later
    be retargeted to SYCL, Triton, or accelerator-native compute.
\end{enumerate}

\textbf{Non-goals.} vkGSplat is not aimed at real-time gaming, VR, or
display-bound interactive workloads. We do not claim to outperform RTX
hardware on its own ground. We claim only that, for the SDG workload,
running graphics on AI compute is viable, economically attractive, and
strategically preferable.

\textbf{Prototype status.} The current implementation is deliberately
CPU-first while preserving the CUDA target. It includes a portable
Wicked Engine Vulkan capture harness, restricted SPIR-V analysis and
C/CUDA translation tests, camera-derived temporal reprojection with
depth and primitive-ID history rejection, an SVGF-style denoising
baseline, a Vulkan-ray-tracing-shaped CPU seed fixture, and a first
native Metal compute backend for Apple Silicon. The
tile-binned 3DGS reference renderer is kept behind an explicit build
flag rather than being part of the default path. On the local MoltenVK
test machine the Wicked harness loads the target scene and reports
SPIR-V shader format, but mesh shaders and Vulkan ray tracing are not
exposed; therefore the current ray-tracing work uses
Vulkan-API-shaped fixtures until a backend with the required extensions
is available. The first such fixture implements a small triangle-scene
seed tracer that emits the same color, depth, NDC-depth, and
primitive-ID buffers expected from a
\texttt{VK\_KHR\_ray\_tracing\_pipeline} capture, then feeds those
buffers through temporal reprojection and denoising. This staging keeps
correctness testable on CPUs while the data layouts and algorithms
remain tile-friendly for CUDA.

% =====================================================================
\section{Background}\label{sec:bg}
% =====================================================================

\subsection{Synthetic data generation for robotics}

The use of simulated environments to train physical agents predates deep
learning, but its modern form was crystallized by domain
randomization~\cite{tobin2017domain} and by simulation platforms such as
Habitat~\cite{savva2019habitat}, ThreeDWorld~\cite{gan2020threedworld},
ProcTHOR~\cite{deitke2022procthor}, and NVIDIA Isaac~Sim. Recent SDG
toolkits like BlenderProc~\cite{denninger2019blenderproc} and
Kubric~\cite{greff2022kubric} provide higher-level scripting for
photorealistic image and video generation with controlled randomization
of geometry, materials, and lighting. NVIDIA Cosmos~\cite{nvidia2024cosmos}
extends this trajectory by combining classical rendering with generative
priors. Across these systems, the rendering layer is the single largest
silicon dependency.

\subsection{Software Vulkan implementations}

Two production software-Vulkan implementations exist.
\textbf{SwiftShader}~\cite{swiftshader}, originally a software renderer
project at TransGaming and now maintained by Google, provides a portable,
LLVM-JIT-based Vulkan~1.3 implementation that runs on CPUs (it is shipped
in Chromium for fallback rendering). \textbf{Mesa Lavapipe}~\cite{mesa-lavapipe}
is the Vulkan front-end on top of Mesa's \texttt{llvmpipe} CPU rasterizer.
Both are correctness-first and CPU-targeted; their performance on GPUs is
either non-existent (Lavapipe is CPU-only) or limited (SwiftShader has
experimental Reactor backends but is not optimized for AI-accelerator
throughput). Neither targets the SDG workload class.

\subsection{Compute-only graphics on GPUs}

There is a long line of research in software rendering on
general-purpose GPU compute. Laine and
Karras~\cite{laine2011software} showed that a fully software rasterizer
on CUDA could approach hardware rasterizer throughput for many workloads,
particularly when rasterization was not the bottleneck.  The same
lineage survives in \textsc{nvdiffrast}~\cite{laine2020nvdiffrast}, which
packages CUDA-accelerated rasterization, interpolation, texturing, and
antialiasing as low-level differentiable rendering primitives.
For ray tracing, Aila and Laine~\cite{aila2009understanding} established
the architectural patterns that would later become OptiX~\cite{parker2010optix}
and Embree~\cite{wald2014embree}. Path tracers including
Mitsuba~3~\cite{jakob2022mitsuba3} (built on Dr.Jit~\cite{jakob2022drjit})
and Falcor~\cite{benty2020falcor} demonstrate that high-quality,
differentiable rendering can be expressed entirely in compute.
Meanwhile, 3D Gaussian Splatting~\cite{kerbl20233dgs} and its
descendants~\cite{yu2024mip,huang20242dgs,ye2024gsplat} have established a new class
of renderer that is compute-native by construction.

The intellectual point is simple: every major step of a modern graphics
pipeline has a published, competitive software-on-compute implementation.
\textbf{What does not exist is a unified Vulkan front-end that exposes
these implementations through a standard API to existing graphics
applications.} That is the gap vkGSplat targets.

% =====================================================================
\section{Modern graphics as acquisition and reconstruction}
\label{sec:acqrec}
% =====================================================================

The standard account of GPU evolution is that fixed-function graphics
hardware became programmable shaders, programmable shaders became
general-purpose compute, and GPUs became a shared substrate for
graphics, scientific computing, and AI. Dally, Keckler, and Kirk tell
this hardware story directly: the GPU began as a dedicated real-time
graphics processor, then accumulated programmability, floating-point
performance, resilience, and AI-scale throughput as new users mapped
their workloads onto it~\cite{dally2021gpu}. The complementary software
story is just as important for vkGSplat: \emph{graphics itself changed
shape}. Rendering moved from a single pipeline traversal, to a
multi-pass software pipeline, to an acquisition--reconstruction
architecture.

\begin{figure*}[t]
\centering
\resizebox{\textwidth}{!}{%
\begin{tikzpicture}[
    font=\small,
    >=Latex,
    era/.style={draw, rounded corners=2pt, very thick, align=center,
        minimum width=4.15cm, minimum height=0.95cm, fill=plotblue!10},
    stage/.style={draw, rounded corners=2pt, align=center,
        minimum width=3.65cm, minimum height=0.74cm, fill=white},
    source/.style={draw, rounded corners=2pt, very thick, align=center,
        minimum width=3.7cm, minimum height=0.9cm, fill=plotcyan!12},
    evidence/.style={draw, rounded corners=2pt, very thick, align=center,
        text width=3.35cm, minimum width=3.65cm, minimum height=1.38cm,
        fill=plotyellow!18},
    recon/.style={draw, rounded corners=2pt, very thick, align=center,
        text width=3.35cm, minimum width=3.65cm, minimum height=1.38cm,
        fill=plotgreen!12},
    arrow/.style={-Latex, thick}
]
% top row: historical rendering eras
\node[era] (era1) at (0,3.15) {\textbf{Era 1}\\fixed-function pipeline};
\node[stage] (era1a) at (0,1.9) {geometry};
\node[stage] (era1b) at (0,0.95) {transform / raster};
\node[stage] (era1c) at (0,0.0) {shade / framebuffer};
\draw[arrow] (era1a) -- (era1b);
\draw[arrow] (era1b) -- (era1c);

\node[era, fill=plotyellow!16] (era2) at (5.15,3.15) {\textbf{Era 2}\\multi-pass software};
\node[stage] (era2a) at (5.15,1.9) {shadow map};
\node[stage] (era2b) at (5.15,0.95) {G-buffer / visibility};
\node[stage] (era2c) at (5.15,0.0) {lighting / denoise / upscale};
\draw[arrow] (era2a) -- (era2b);
\draw[arrow] (era2b) -- (era2c);

\node[era, fill=plotgreen!14] (era3) at (10.3,3.15) {\textbf{Era 3}\\AI inference pipeline};
\node[stage] (era3a) at (10.3,1.9) {thin acquisition};
\node[stage, fill=plotgreen!8] (era3b) at (10.3,0.95) {learned reconstruction};
\node[stage, fill=plotgreen!8] (era3c) at (10.3,0.0) {frame / world synthesis};
\draw[arrow] (era3a) -- (era3b);
\draw[arrow] (era3b) -- (era3c);

\draw[-Latex, very thick] (1.85,0.95) -- node[above, align=center, font=\footnotesize]
    {fragment into\\partial passes} (3.3,0.95);
\draw[-Latex, very thick] (7.0,0.95) -- node[above, align=center, font=\footnotesize]
    {render less,\\infer more} (8.45,0.95);

% bottom row: acquisition equivalence
\node[source] (real) at (0,-2.75) {\textbf{Physical acquisition}\\camera / sensors};
\node[stage] (photons) at (0,-4.05) {photons, depth, IMU, flow};
\draw[arrow] (real) -- (photons);

\node[source, fill=plotblue!10] (synthetic) at (4.25,-2.75) {\textbf{Synthetic acquisition}\\renderer / simulated camera};
\node[stage] (primary) at (4.25,-4.05) {primary visibility};
\node[stage] (shadow) at (4.25,-5.45) {shadow / light visibility};
\node[stage] (secondary) at (4.25,-6.45) {secondary visibility rays};
\draw[arrow] (synthetic) -- (primary);
\draw[arrow] (primary) -- (shadow);
\draw[arrow] (shadow) -- (secondary);

\node[evidence] (evidence) at (8.8,-5.1) {\textbf{Structured evidence}\\depth, normals, IDs, motion,\\sparse radiance, history};
\node[recon] (recon) at (13.2,-5.1) {\textbf{Reconstruction}\\denoise, TAA, upscale,\\frame generation, neural fields};
\draw[-Latex, very thick] (photons.east) -- ++(0.45,0) |- ([yshift=0.28cm]evidence.west);
\draw[-Latex, very thick] (secondary.east) -- ++(0.45,0) |- ([yshift=-0.28cm]evidence.west);
\draw[-Latex, very thick] (evidence) -- (recon);
\node[font=\footnotesize, align=center] at (8.8,-7.45)
    {same reconstruction interface; different acquisition source};
\end{tikzpicture}%
}
\caption{The graphics model that vkGSplat relies on. The upper row shows
the historical shift from a single fixed-function pipeline, to a graph
of partial software passes, to an AI inference pipeline. The lower row
shows the acquisition--reconstruction split: real-world images and
synthetic images differ mainly in the source of acquisition. A real
camera captures physical evidence; a renderer captures structured
evidence from a simulated world through primary visibility, shadow or
light visibility, and secondary visibility. Both then feed
reconstruction.}
\label{fig:graphics-evolution}
\end{figure*}

\subsection{From one pipeline to partial pipelines}

Classical rendering was organized around a relatively direct flow:
geometry entered the pipeline, fixed-function units transformed and
rasterized it, shaders ran at programmable stages, and pixels were
written to a framebuffer~\cite{akenine2018}. Even after vertex and
pixel shaders became programmable, the larger execution model remained
inherited from fixed-function graphics: the shader was programmable,
but the pipeline was not.

Deferred shading changed the abstraction. It separated visibility from
shading by first projecting scene information into screen-space buffers
--- depth, normals, material parameters, motion vectors, and primitive
identity --- and only then shading or filtering those buffers as later
passes~\cite{akenine2018}. Once geometry has been regularized into
screen-space arrays, the problem becomes software-defined: arrays can
be consumed by compute shaders, denoisers, temporal filters, neural
upscalers, and frame-generation models. Modern rendering is therefore
not one pipeline traversal. It is a graph of partial pipelines:
shadow-map passes that need only depth, G-buffer passes that need only
attributes, ray-tracing passes that need sparse secondary visibility,
denoisers that need noisy radiance plus guides, and upscalers that need
a low-resolution frame plus history.

This matters because partial pipelines are much easier to reinterpret
as compute kernels. A shadow pass is a depth-acquisition kernel. A
primary-visibility pass is a camera-acquisition kernel. A
secondary-visibility pass is a sparse sampling kernel. A denoiser or TAA
pass is a reconstruction kernel. Vulkan remains the application-level
description of the work, but the implementation no longer has to be a
hardware raster pipeline.

\subsection{Graphics shaders as restricted compute}

The word ``graphics'' can obscure where modern rendering time actually goes.
On contemporary GPUs, the large programmable resource is the shader/SIMT
execution fabric plus memory hierarchy; fixed-function raster, texture, ROP,
and RT blocks are specialized accelerators around that fabric, not a separate
universe of computation. Exact die-area ratios are vendor-specific and usually
not public, but architecturally the trend is clear: graphics pipelines have
accumulated more programmable shader work, more full-screen passes, more
temporal history, and more tensor-side reconstruction. The classic
fixed-function stages increasingly acquire and organize evidence; the expensive
value-building stages are kernels over buffers.

In this sense, a graphics shader is already a restricted compute kernel. A
vertex, mesh, fragment, ray-generation, closest-hit, or miss shader is ordinary
program code launched over a pipeline-defined domain: vertices, primitives,
fragments, rays, or samples. Its memory interface and scheduling are constrained
by the graphics pipeline, and historically many writes are mediated through
attachments, interpolation, blending, depth/stencil, or descriptor resources.
A compute shader exposes the next degree of freedom: explicit workgroups,
shared memory, storage buffers, storage images, arbitrary load/store, and
user-defined data layouts. Many modern graphics effects exploit exactly that
freedom: tiled lighting, SSAO, SSR, denoisers, TAA, temporal upscalers, particle
updates, culling, skinning, visibility compaction, and neural reconstruction
are already screen-space or buffer-space compute passes.

vkGSplat pushes the same migration one step further. It does not merely replace
a fragment pass with a compute pass over the same screen-space image. It asks
whether the renderer's evidence should live in richer address spaces:
world-space Gaussian state, 4D motion state, path-space radiance caches,
material and primitive epochs, tile pages, and uncertainty buffers. Screen-space
compute treats the image as the main memory of the renderer. vkGSplat treats the
image as one projection of a persistent physical-evidence state. That is the
stronger meaning of ``compute-first'': the software stack is no longer limited
to framebuffer-sized arrays or pipeline-defined shader invocations, even though
it can still ingest ordinary Vulkan graphics output as an acquisition source.

\subsection{Real and synthetic images share the same reconstruction
contract}

In physical vision, acquisition is sensor capture: photons from the
real world become a noisy, finite-resolution image, often accompanied
by depth, IMU, optical flow, or multi-camera context. In synthetic
vision, acquisition is the early part of the renderer. The renderer is
also a camera, but it observes a simulated world: primary visibility
determines what the virtual camera sees; shadow-map or other
light-visibility passes determine which emitters can contribute;
secondary-visibility rays sample reflection, refraction, indirect
illumination, and occlusion; depth, normals, motion vectors, material
IDs, primitive IDs, and sparse radiance record the evidence.

The source-specific difference is therefore not that one pipeline
produces ``images'' and the other produces ``graphics.'' Both are
acquisition pipelines followed by reconstruction. The only difference is
where the acquisition comes from: a physical camera observing the real
world, a hardware Vulkan renderer observing a simulated world through graphics
units, an offline path tracer, or a software renderer observing the same
simulated world through CUDA kernels. This reframing is central to vkGSplat. If
synthetic acquisition can be represented by a common seed-buffer contract, then
the reconstruction path --- temporal reprojection, history rejection, denoising,
super-resolution, and eventually learned reconstruction --- can consume either
hardware-produced or software-produced evidence. The CUDA software path is the
unique vkGSplat bet, but it is not the only valid acquisition source.

\subsection{Acquisition backends: hardware Vulkan and CUDA software}
\label{sec:acquisition-backends}

The acquisition--reconstruction philosophy is deliberately backend-neutral. A
Vulkan graphics pipeline running on RTX-class hardware is a first-class way to
produce physical or synthetic visual evidence. In many deployments it is the
right answer: it has mature drivers, high raster and ray-tracing throughput,
hardware texture filtering, mature debugging tools, and broad application
compatibility. A vkGSplat system should therefore be able to ingest hardware
Vulkan outputs directly: color, depth, motion vectors, primitive IDs, material
IDs, G-buffer attributes, sparse radiance buffers, and sample statistics when the
renderer exposes them.

The CUDA software rasterizer is a different point in the design space, not a
claim that graphics hardware is obsolete. Its benefits are strongest when the
consumer is an AI pipeline rather than a display pipeline:
\begin{enumerate}
  \item \textbf{Hardware access and fleet utilization.} H100/B200-class
        accelerators may be available where RTX/L40-class graphics GPUs are not.
  \item \textbf{Co-location with reconstruction and training.} Seed generation,
        Gaussian-state updates, denoising, neural resolving, and policy or
        world-model training can stay in CUDA/PyTorch memory without repeated
        graphics-queue interop or host readback.
  \item \textbf{Programmable evidence.} The renderer can emit exactly the
        metadata the reconstructor needs: path-depth statistics, variance,
        confidence, primitive/material IDs, visibility gates, and per-sample
        provenance.
  \item \textbf{Deterministic offline scheduling.} SDG can batch cameras,
        worlds, and samples for throughput, reproducibility, and adaptive oracle
        refresh rather than optimizing a 16-ms interactive frame.
  \item \textbf{Portability to non-graphics accelerators.} A tile- and
        tensor-friendly software contract can later be lowered to SYCL, Triton,
        XLA, or Neuron-style backends where no Vulkan graphics queue exists.
\end{enumerate}

This makes the comparison more precise. Hardware Vulkan is the baseline and often
the oracle for conventional rendering performance. CUDA software acquisition is
valuable when the marginal cost of leaving the AI accelerator, losing custom
metadata, or provisioning a separate graphics fleet exceeds the efficiency of
fixed-function graphics hardware. The paper's system claim is therefore not
``software rasterization always wins,'' but ``Physical AI data generation should
expose a common acquisition contract, and vkGSplat tests the software-CUDA point
on that contract.''

\subsection{Physical simulation and sensor modeling are orthogonal}
\label{sec:physical-sim-sensors}

For Physical AI, the acquisition--reconstruction split has a sharper
meaning: physical realism must be produced at acquisition time. A
Gaussian state can preserve, fuse, denoise, and reconstruct evidence,
but it cannot create physical truth that was never simulated or
measured. Contact impulses, joint torques, IMU bias, rolling-shutter
readout, LiDAR dropout, multipath, exposure, lens distortion, and
material-dependent sensor response are not consequences of 3D Gaussian
Splatting. They belong to the physical simulator and the sensor model.

This makes physical simulation and sensor modeling orthogonal to the
GSplat reconstruction backend. Let $S_t$ denote the physical simulation
state: articulated poses, velocities, contact state, mass, friction,
lighting, materials, and environment state. Let $Q_s$ denote the
calibrated acquisition model for sensor $s$, with calibration $C_s$ and
noise or corruption variables $\eta_s$. The sensor evidence is
\begin{equation}
  A_{s,t} = Q_s(S_t, C_s, \eta_s),
  \label{eq:sensor-acquisition}
\end{equation}
while the reconstruction backend consumes that evidence through a
persistent state update
\begin{equation}
  \mathcal{G}_t = U(\mathcal{G}_{t-1}, \{A_{s,t}\}_s).
  \label{eq:gs-reconstruction-update}
\end{equation}
The former equation is about physical and sensor correctness; the latter
is about efficient representation, temporal reuse, uncertainty tracking,
and reconstruction. They should share timestamps, coordinate frames,
calibration, primitive or semantic IDs, and confidence estimates, but
they should not be collapsed into one algorithmic claim.

This is consistent with the organization of robotics simulators. Isaac
Sim is presented as a robotics simulation and synthetic-data platform,
while its physics-based sensors are documented separately from rendering:
they access physics properties such as mass and velocity after the
render step and can be paired with noise models~\cite{nvidia-isaacsim,
nvidia-isaac-physics-sensors}. The same separation should hold in
vkGSplat. RGB, depth, LiDAR, radar, event cameras, IMU, joint, and
contact sensors each require their own $Q_s$. GSplat reconstruction can
then act as a fast measured-state cache for visual and geometric sensor
evidence, with uncertainty gates that fall back to exact physics,
ray/path tracing, or a reference sensor simulator when the Gaussian
state is underspecified.

The design consequence is simple. The master vkGSplat architecture owns
three separable layers:
\begin{enumerate}
  \item \textbf{Physical simulation:} dynamics, contacts, materials,
        lights, robot state, and scene evolution.
  \item \textbf{Sensor acquisition:} camera optics, exposure, rolling
        shutter, LiDAR/radar scan timing, noise, dropout, quantization,
        calibration, and sensor-specific failure modes.
  \item \textbf{GSplat reconstruction:} persistent Gaussian state,
        confidence, variance, temporal fusion, denoising, TAA,
        super-resolution, and optional learned resolving.
\end{enumerate}
A renderer can be compute-first without pretending these layers are the
same. In this paper, ``compute-first Vulkan path'' means a compute-first
implementation of Vulkan's visual acquisition/rendering contract: visibility,
G-buffer evidence, sparse radiance, texture/material evidence, and image-space
outputs. It does not mean Vulkan becomes the contract for rigid-body dynamics,
contact, IMU, LiDAR, radar, or all sensor semantics. vkGSplat's core claim is
that the reconstruction layer, and much of the structured visual acquisition
layer, can run on tensor-rich compute hardware. It is not that a Gaussian
reconstructor replaces physical simulation or calibrated sensor modeling.

\subsection{RT cores, tensor cores, and the new split}

Once the acquisition--reconstruction split is visible, RT cores and
tensor cores no longer look like unrelated add-ons. RT cores accelerate
bounded acquisition: traversal, intersection, and sparse visibility
sampling. Tensor cores accelerate value-scalable reconstruction:
denoising, upscaling, frame generation, neural radiance caching, and
learned scene synthesis. In the language of the modernized Amdahl
framework~\cite{lu2026amdahl}, acquisition is mostly bounded once the
chosen resolution, sample count, and visibility problem are fixed;
reconstruction is value-scalable because additional model capacity and
inference compute keep improving perceptual quality, consistent with AI
scaling behavior~\cite{kaplan2020,lu2026}.

For real-time games, fixed-function acquisition remains valuable because
the frame budget is tight. For SDG, the balance changes. The consumer is
usually a downstream model, not a human staring at a 16-ms deadline, and
the workload is priced in images per dollar or joule. Hardware Vulkan remains a
strong acquisition baseline, especially when RTX/L40-class GPUs are already in
the loop. vkGSplat explores the other branch: keep the Vulkan description and
seed-buffer contract, but implement acquisition and reconstruction as
tile-friendly compute when AI-hardware co-location and programmability dominate
the cost model.

\subsection{The DLSS precedent, and why vkGSplat replaces its role with
integrated 3DGS}
\label{sec:dlss}

The strongest evidence that graphics workloads can move off
graphics-specific silicon is that they already have, partially, inside
NVIDIA's own products. DLSS-class pipelines reduce the renderer's job
from producing the final display image to producing a sparse seed:
sub-native resolution color, depth, motion, and temporally jittered
samples. Neural reconstruction then performs super-resolution,
anti-aliasing, denoising, and, in DLSS~3, frame
generation~\cite{nvidia-dlss,liu2020nsrr,edelsten2019dlss,nvidia-dlss3}.
In other words, DLSS is not primarily a rasterizer or RT-core story. It is a
compute reconstruction story that has been productized as part of the graphics
stack. vkGSplat's claim is that SDG needs that same reconstruction role, but
with an open, explicit, device-resident 3DGS/radiance state instead of a closed
image-space model packaged with graphics hardware.
Ray-traced global illumination follows the same pattern:
single-sample-per-pixel or low-sample path tracing produces sparse
radiance, then spatiotemporal filters and denoisers reconstruct the
image~\cite{schied2017svgf,chaitanya2017interactive,nvidia-nrd,bitterli2020restir}.
NVIDIA's DLSS Ray Reconstruction makes this boundary explicit: ray samples are
still fired into the scene to acquire representative lighting, reflectivity,
and shadowing, while an AI network replaces hand-tuned denoisers and generates
higher-quality pixels between sampled rays~\cite{nvidia-dlss35rr}. In other
words, the physical grounding comes from the low-sample path-traced acquisition;
the neural stage reconstructs the missing samples. This is exactly the
principle vkGSplat should adopt for Physical AI visual realism.

The caveat is important. Low-sample path tracing plus AI denoising can preserve
physical plausibility and much of the realism of path tracing, but it is not an
unbiased physical measurement. The denoiser may suppress rare paths, invent
smooth structure, or fail under distribution shift. For SDG and robotics, the
right contract is therefore not ``AI makes it physical,'' but ``physical rays
acquire sparse truth; reconstruction fills the gaps; uncertainty gates decide
when more oracle samples or a reference sensor path are required.''

The sample budget and the path-depth budget should also be separated. Reducing
samples per pixel lowers Monte Carlo acquisition cost while preserving the same
transport integral in expectation. Reducing the maximum number of bounces changes
the physical model itself: indirect illumination, refraction chains, glossy
interreflection, and caustics can disappear before reconstruction ever sees
support for them. A vkGSplat acquisition pass should therefore prefer low
samples per pixel, lower seed resolution, adaptive sampling, and Russian
roulette over a hard shallow-bounce cutoff. Each seed should carry enough
metadata for the reconstructor to know how much physics it has actually seen:
\texttt{spp}, requested maximum depth, realized path-depth moments, Russian
roulette survival statistics, radiance variance, and confidence. A Gaussian or
neural resolver can amortize sparse long-path evidence over views and time, but
when those statistics say that a region has never observed the relevant path
class, the system should request more oracle samples instead of silently
hallucinating realism.

vkGSplat does not need to clone DLSS; it needs to replace DLSS's architectural
role for SDG. DLSS proves that the valuable part of modern rendering has moved
toward compute-side reconstruction, but DLSS itself is closed, display-oriented,
and coupled to a product stack optimized for games. vkGSplat's replacement is
integrated 3DGS: persistent Gaussian/radiance state that stores geometry,
radiance, motion, confidence, variance, and validity epochs, then serves
denoising, TAA, super-resolution, frame interpolation, radiance-cache reuse, and
optional learned residuals. The vkGSplat question is therefore narrower and
more actionable: if the seed is already sparse, noisy, and reconstructed anyway,
can software rendering produce that seed on AI compute cheaply enough, and can
integrated 3DGS reconstruct it well enough, for SDG? Published software
rasterization~\cite{laine2011software}, GPU ray traversal
~\cite{aila2009understanding}, neural supersampling~\cite{xiao2020}, and
compute-native 3DGS~\cite{kerbl20233dgs} make the answer plausible. vkGSplat is
the engineering experiment that turns that plausibility into a
Vulkan-compatible stack.

% =====================================================================
\section{Modernized Amdahl's Law: why SDG moves value to compute}
\label{sec:amdahl-main}
% =====================================================================

The companion Amdahl paper reframes the old serial-versus-parallel
question as a hardware-allocation question: with a constrained silicon
or cluster budget, how much should be committed to narrowly specialized
logic, and how much should remain programmable while the workload itself
keeps changing?  The detailed model is in
Appendix~\ref{app:amdahl}; the architectural takeaway is what matters
here.  A fixed-function block is attractive only while it delivers a
large relative efficiency advantage on a stable, value-bounded share of
the workload.  As more of the delivered value moves into stages that
keep improving with model capacity, inference compute, temporal history,
and software updates, the efficiency bar for narrow specialization rises.

\noindent\textbf{Main result: Modernized Amdahl's Law for vkGSplat.}
The decision is not whether fixed-function graphics can accelerate one bounded
stage. The decision is whether that bounded acquisition share is large enough
to justify a separate graphics allocation once reconstruction, world modeling,
and training become the value-scalable share. Let $x$ be the fraction of the
hardware or cluster budget assigned to specialized graphics logic, let $R$ be
the relative efficiency advantage of that logic on the bounded acquisition
share, and let $s$ be the value-scalable share whose utility keeps improving on
programmable compute. The central allocation objective is
\begin{equation}
\boxed{
  T(R,s,x)=
  \frac{1-s}{1+(R-1)x} + \frac{s}{1-x}
}
\label{eq:main-modern-amdahl-objective}
\end{equation}
where lower $T$ means lower normalized cost per useful SDG sample. In
plain-text form for downstream summarizers, the objective name is
\mbox{\texttt{T(R, s, x)}} and the equation reads
\texttt{T = (1-s)/(1+(R-1)x) + s/(1-x)}. The fixed-function allocation is
justified only if
\begin{equation}
\boxed{
  R > R_c(s)=\frac{1}{1-s}
}
\label{eq:main-amdahl-decision-rule}
\end{equation}
which means the required efficiency advantage diverges as $s\rightarrow 1$.
This is the modernized Amdahl's Law argument for vkGSplat: once DLSS-like
reconstruction, integrated 3DGS state, world models, and training dominate the
useful value of SDG, programmable AI compute becomes the natural substrate.

The same objective also gives the critical scalable fraction for a fixed
specialized efficiency ratio:
\begin{equation}
\boxed{
\begin{aligned}
  R_c(s) &= \frac{1}{1-s},\\
  s_c(R) &= 1-\frac{1}{R}.
\end{aligned}}
\label{eq:main-modern-amdahl}
\end{equation}
The objective $T(R,s,x)$ is not meant to predict a single frame time. It is a
first-order allocation model for hardware and cluster budgets that ignores
second-order effects such as
interconnect topology, utilization policy, and data movement except insofar as
they change the effective $R$ or the measured scalable share $s$. In vkGSplat,
$s$ should be estimated empirically by ablating acquisition, reconstruction,
training, and world-model stages on a fixed workload. The interpretation is
deliberately sharp: as reconstruction and learned world modeling push $s$ toward
one, the efficiency ratio required to justify a separate fixed-function
allocation diverges. This is the modernized Amdahl argument for making SDG a
compute-first AI datacenter workload rather than a separate graphics-cluster
workload.

This is exactly the structure of modern rendering.  Visibility, shadow
tests, depth, motion vectors, and sparse radiance are bounded acquisition
signals once the camera, resolution, and sample budget are chosen.  The
stage that still scales in value is reconstruction: denoising,
anti-aliasing, super-resolution, frame synthesis, neural radiance
caching, and eventually 3DGS or world-model reconstruction. The same
principle appears in any low-sample Monte Carlo renderer: the seed is a
sparse, noisy acquisition signal, while the denoising or neural
reconstruction stage recovers the image that downstream consumers
actually use.

For real-time games, specialized acquisition hardware can remain the
right trade because the 16-ms display budget makes fixed-function
latency valuable.  For SDG, the value function changes.  The consumer is
a training pipeline, not a human display loop, and the economic objective
is images per dollar or joule.  Under the modernized Amdahl view,
vkGSplat deliberately puts the bounded acquisition stages into software
kernels and reserves the programmable substrate for the part of the
pipeline that will keep changing: reconstruction, denoising, temporal
reuse, and learned scene representations.

% =====================================================================
\section{Why SDG is the right workload}\label{sec:workload}
% =====================================================================

\subsection{Amortized world building before simulation}
\label{sec:world-building-lifecycle}

The most important difference between SDG and gaming is not just that SDG can
tolerate higher per-image latency. It is that SDG can separate
\emph{world-building time} from \emph{simulation and acquisition time}. A game
renderer must satisfy the next display frame under a hard latency budget. A SDG
pipeline can spend a large offline budget establishing the world before it runs
many camera trajectories, robot rollouts, randomized episodes, or policy
training batches.

The canonical vkGSplat lifecycle is therefore:
\[
\text{build} \rightarrow \text{simulate} \rightarrow \text{update}
\rightarrow \text{acquire} \rightarrow \text{reconstruct}.
\]
The operational slogan is: \emph{build once, simulate many, invalidate
precisely, acquire sparsely, reconstruct densely}. World building creates an
expensive reusable base; simulation advances the changing physical state; update
applies transforms and marks invalid cache regions; acquisition resamples only
changed or uncertain evidence; reconstruction fuses that evidence into images,
labels, Gaussian state, radiance caches, and sensor outputs.

\begin{figure}[t]
\centering
\resizebox{\textwidth}{!}{%
\begin{tikzpicture}[
  phase/.style={draw, rounded corners=2pt, very thick, align=center,
    text width=2.55cm, minimum height=1.08cm, font=\scriptsize},
  statebox/.style={draw, dashed, rounded corners=2pt, align=center,
    text width=11.9cm, minimum height=1.12cm, font=\scriptsize,
    fill=plotgreen!8},
  edge/.style={-{Latex[length=2mm]}, thick},
  feedback/.style={-{Latex[length=2mm]}, thick, dashed},
  exchange/.style={Latex-Latex, thick, dashed},
  note/.style={align=center, font=\scriptsize, inner sep=1pt}
]
\node[phase, fill=plotblue!10] (build) at (0,0)
  {\textbf{Build}\\static world $\mathcal{W}_0$\\geometry, materials\\BVH, 3DGS, caches};
\node[phase, fill=plotcyan!10] (simulate) at (3.15,0)
  {\textbf{Simulate}\\dynamic state $\mathcal{D}_t$\\robots, contacts\\objects, lights};
\node[phase, fill=plotyellow!15] (update) at (6.30,0)
  {\textbf{Update}\\transforms, epochs\\dirty regions\\confidence decay};
\node[phase, fill=plotred!10] (acquire) at (9.45,0)
  {\textbf{Acquire}\\Vulkan, RT, sensors\\seed buffers\\sparse truth};
\node[phase, fill=plotgreen!12] (reconstruct) at (12.60,0)
  {\textbf{Reconstruct}\\Gaussian/radiance state\\denoise, TAA, SR\\frame interpolation};

\node[statebox] (state) at (6.30,-1.85)
  {\textbf{Persistent explicit state}\\
   $\mathcal{W}_0$ static base \quad $\mathcal{D}_t$ dynamic overlay \quad $\mathcal{G}_t$ Gaussian/radiance state\\
   caches, sensor history, variance, confidence, and validity epochs};
\node[phase, fill=white] (oracle) at (1.65,-3.70)
  {\textbf{Oracle refresh}\\extra rays/path depth\\sensor simulator\\physics checks};
\node[phase, fill=white] (consumer) at (10.95,-3.70)
  {\textbf{Consumers}\\images, labels\\sensor streams\\training rollouts};

\coordinate (stateBuild) at ($(state.north west)+(1.15,0)$);
\coordinate (stateUpdate) at (state.north);
\coordinate (stateRecon) at ($(state.north east)+(-1.15,0)$);
\coordinate (stateQuery) at ($(state.north east)+(-3.15,0)$);

\draw[edge] (build) -- (simulate);
\draw[edge] (simulate) -- (update);
\draw[edge] (update) -- (acquire);
\draw[edge] (acquire) -- (reconstruct);
\draw[edge] (build.south) -- (stateBuild);
\draw[edge] (update.south) -- (stateUpdate);
\draw[edge] (reconstruct.south) -- (stateRecon);
\draw[edge] (stateQuery) -- (acquire.south);

\draw[edge] (reconstruct.south east) to[out=-65,in=65] (consumer.north east);
\draw[feedback] (consumer.north) -- (state.south east);
\draw[exchange] (state.south west) -- (oracle.north);
\end{tikzpicture}%
}
\caption{Whole SDG lifecycle for vkGSplat. Expensive world building creates a
reusable static base; simulation produces dynamic overlays; update/invalidation
marks what changed; acquisition collects sparse physical evidence through
hardware Vulkan, RT/path tracing, or sensor models; reconstruction turns that
evidence into dense images, labels, sensor streams, Gaussian state, and radiance
caches. The consumer is usually the next robot or world model, whose rollout
distribution feeds back into simulation. Uncertainty and validity epochs decide
whether to reuse state, locally reacquire evidence, or return to the oracle or
rebuild stage.}
\label{fig:sdg-lifecycle}
\end{figure}

\subsection{Closed-loop model training}
\label{sec:closed-loop-training}

The lifecycle in Fig.~\ref{fig:sdg-lifecycle} is also a training loop. Let
$M_{\theta_k}$ be the current robot model: a perception model, policy, planner,
world model, or a coupled stack of these components. At iteration $k$, that
model acts in the simulated or reconstructed world, producing trajectories
$\tau_k$, task outcomes, uncertainty signals, and failure cases. vkGSplat then
uses those signals to choose what to acquire, reconstruct, label, and refresh:
\begin{equation}
\boxed{
\begin{aligned}
  \tau_k
    &\sim \mathrm{Rollout}(M_{\theta_k}, S_t),\\
  \mathcal{D}^{\mathrm{syn}}_k
    &= \mathrm{AcquireReconstructLabel}(\tau_k, \mathcal{G}_t, Q_s),\\
  \theta_{k+1}
    &= \mathrm{Train}\!\left(\theta_k,
       \mathcal{D}^{\mathrm{real}} \cup \mathcal{D}^{\mathrm{syn}}_k,
       \mathcal{L}_{\mathrm{task}}\right).
\end{aligned}}
\label{eq:closed-loop-sdg-training}
\end{equation}
The important point is that $\mathcal{D}^{\mathrm{syn}}_k$ is not sampled from a
fixed offline distribution. It is shaped by the current model's behavior: where
the robot collides, hesitates, misperceives, loses contact, fails grasp
closure, violates a safety margin, or produces high epistemic uncertainty. This
turns SDG into active data generation. The model asks the world for the data it
needs next; acquisition and reconstruction make that data cheap enough to
generate at scale; training changes the model that will ask the next question.

We should therefore model a generated world as an amortized static base plus
dynamic overlays:
\begin{equation}
  S_t = \mathcal{W}_0 \oplus \mathcal{D}_t,
  \label{eq:world-plus-dynamics}
\end{equation}
where $\mathcal{W}_0$ contains static geometry, materials, textures, calibrated
lighting priors, BVHs, mesh or Gaussian maps, precomputed visibility, and
high-quality radiance or light-transport caches. $\mathcal{D}_t$ contains the
state that changes during simulation: robot joints, contacts, movable objects,
deformables, particles, dynamic lights, animated materials, and environment
edits. Acquisition then measures
\begin{equation}
  A_{s,t} = Q_s(\mathcal{W}_0 \oplus \mathcal{D}_t, C_s, \eta_s),
  \label{eq:world-lifecycle-acquisition}
\end{equation}
and reconstruction updates explicit state as in
Eq.~\ref{eq:gs-reconstruction-update}. The key is that $\mathcal{W}_0$ can be
expensive, because it is reused across many $t$, sensors, camera paths,
randomized object placements, and downstream training jobs.

This changes the role of path tracing and Gaussian state. In the
world-building phase, vkGSplat can afford high-spp, high-bounce reference
passes, inverse-rendered material estimation, dense static BVHs, static 3DGS
capture, radiance-cache warmup, and sensor-calibration validation. In the
simulation phase, it should not repeat that work blindly. It should acquire the
delta caused by $\mathcal{D}_t$, reuse the static evidence where valid, and ask
the oracle for additional samples only in regions whose visibility, lighting,
material, or sensor state has changed.

\begin{table*}[t]
\centering
\footnotesize
\caption{SDG lifecycle. Unlike games, SDG can amortize expensive
world-building over many later simulation and acquisition episodes.}
\label{tab:sdg-lifecycle}
\begin{tabular}{p{0.18\textwidth}p{0.22\textwidth}p{0.27\textwidth}p{0.23\textwidth}}
\toprule
\textbf{Phase} & \textbf{Budget} & \textbf{Products} & \textbf{Main risk} \\
\midrule
World building & Seconds to hours per world & Static geometry, materials, textures, BVH/TLAS base, light priors, Gaussian maps, high-bounce radiance or transport caches & Baking final RGB that becomes invalid under later dynamics \\
Simulation & Step-level physics budget & Robot state, contacts, movable objects, deformables, dynamic lights, material edits, semantic labels & Treating dynamic changes as only visual changes rather than physical state changes \\
Update / invalidation & Per step or event-driven & Dynamic transforms, dirty camera/shadow/radiance masks, cache epoch updates, confidence decay & Missing a dependency and reusing stale evidence \\
Acquisition & Per frame or adaptive burst & Seed buffers: color, depth, normals, primitive/material IDs, motion, sparse radiance, variance, path-depth statistics & Under-sampling changed visibility or changed lighting \\
Reconstruction & Batched CUDA/PyTorch budget & Gaussian/radiance cache update, denoising, TAA, SR, frame interpolation, neural residuals & Reusing stale static evidence after invalidation events \\
Oracle refresh & Adaptive, uncertainty-triggered & Extra ray/path samples, sensor-simulator calls, high-fidelity physics or material evaluation & Failing to refresh rare but task-critical events \\
\bottomrule
\end{tabular}
\end{table*}

Dynamic lighting is the caveat that makes this lifecycle precise. If the static
world stores only final RGB radiance, it is tied to the lights, materials, and
visibility that produced it. Camera motion and static occlusion can be handled
by depth, primitive IDs, motion, and disocclusion gates, but changing an emitter
or material changes the radiometric problem itself. The correct static product
is therefore not always a baked color image. It may need to be a relightable
state: albedo and BRDF parameters, normals, visibility, per-light or
per-environment-map transport bases, spherical-harmonic incident-light
coefficients, path-depth moments, or separate direct and indirect radiance
subslots. The minimum rule is that every radiance cache entry should carry
validity epochs:
\[
  e = (e_{\mathrm{geom}}, e_{\mathrm{mat}}, e_{\mathrm{light}}, e_{\mathrm{vis}}),
\]
and a reconstruction query is high-confidence only when those epochs match the
current dynamic state or when a learned residual has been validated for that
kind of change.

\begin{table*}[t]
\centering
\footnotesize
\caption{Validity of static world-building products under dynamic changes.}
\label{tab:dynamic-validity}
\begin{tabular}{p{0.18\textwidth}p{0.27\textwidth}p{0.43\textwidth}}
\toprule
\textbf{Change} & \textbf{What remains reusable} & \textbf{Required vkGSplat response} \\
\midrule
Camera or robot motion & Static geometry, material, BVH, Gaussian map, static radiance if lighting is unchanged & Reproject using depth, motion, primitive IDs, and visibility gates \\
Dynamic rigid object & Static background state; object-local submap in local coordinates & Update TLAS/object transform, acquire new occlusion and shadow evidence, gate stale background behind object masks \\
Deformable, fluid, particles & Static environment only & Treat as dynamic acquisition; use oracle samples or a specialized simulator/sensor model \\
Dynamic occluder & Material and static lighting priors may remain valid & Refresh direct visibility, shadow visibility, and disoccluded regions; do not trust hidden Gaussian radiance blindly \\
Moving or changing light & Geometry, albedo, BRDF, normals, and transport basis may remain valid; baked RGB does not & Invalidate light epoch, use per-light/SH transport if available, otherwise retrace sparse radiance and update cache \\
Material or texture edit & Geometry, visibility, and light state may remain valid & Invalidate material epoch, recompute BRDF-dependent direct and indirect radiance \\
Emissive object change & Geometry may remain valid, but both light and material epochs change & Treat as dynamic light plus material edit; refresh nearby indirect radiance and shadows \\
\bottomrule
\end{tabular}
\end{table*}

This is where 3DGS should be used carefully. A pure color-only Gaussian scene is
an excellent static visual cache, but it is not a physical lighting model. A
Physical-AI-grade vkGSplat state should move toward \emph{relightable Gaussians}:
geometry, covariance, normal, albedo/material, visibility confidence, motion,
and radiance decomposed into static, direct-light, indirect-light, and learned
residual components. That does not make GSplat a replacement for path tracing or
sensor simulation. It makes GSplat the amortized memory that decides which
parts of the world can be reused and which parts require renewed acquisition.

\subsection{Dynamic changes require invalidation, not full reconstruction}
\label{sec:dynamic-invalidation}

Dynamic objects, shadow changes, occlusion changes, and moving lights should not
force vkGSplat to reconstruct the entire world. They should force a
\emph{dependency-aware partial refresh}. The static world state answers what
exists, where it is, and what its material parameters are. Dynamic acquisition
answers what is visible and how it is illuminated under the current camera,
light, and object state. Confusing these two questions is the failure mode that
turns a useful 3DGS cache into a stale RGB hallucination.

Each Gaussian or radiance-cache entry should therefore carry dependency metadata
in addition to color and covariance:
\[
g_i =
(\mu_i,\Sigma_i,n_i,\rho_i,m_i,L_i,\sigma_i^2,c_i,
 \mathcal{O}_i,\mathcal{L}_i,e_i),
\]
where $\rho_i$ is albedo or material state, $m_i$ is motion state, $L_i$
contains direct, indirect, and residual radiance subslots, $\sigma_i^2$ and
$c_i$ store variance and confidence, $\mathcal{O}_i$ is the set of object or
primitive IDs that can occlude or move the entry, $\mathcal{L}_i$ is the set of
lights or environment components that materially affect it, and
$e_i=(e_{\mathrm{geom}},e_{\mathrm{mat}},e_{\mathrm{light}},e_{\mathrm{vis}})$
stores validity epochs. A camera query, shadow query, or radiance query is
trusted only if its dependency epochs match the current scene state and its
confidence remains above the task threshold.

At simulation step $t$, the physics and scene systems emit change events:
\[
\Delta_t =
\{\Delta_{\mathrm{pose}}, \Delta_{\mathrm{geom}}, \Delta_{\mathrm{mat}},
  \Delta_{\mathrm{light}}, \Delta_{\mathrm{camera}}\}.
\]
vkGSplat maps these events to dirty sets:
\begin{align}
  \mathcal{D}^{cam}_t &= \mathrm{Project}(\Delta_{\mathrm{pose}}
      \cup \Delta_{\mathrm{geom}}, C_t),\\
  \mathcal{D}^{shadow}_t &= \mathrm{ProjectLight}(\Delta_{\mathrm{pose}}
      \cup \Delta_{\mathrm{geom}}, \{L_t\}),\\
  \mathcal{D}^{rad}_t &= \mathrm{Influence}(\Delta_{\mathrm{light}}
      \cup \Delta_{\mathrm{mat}} \cup \mathcal{D}^{shadow}_t).
\end{align}
Only these regions need new evidence. A moved object updates its transform, TLAS
instance, object-local Gaussian transform, and camera/depth occlusion mask. The
background it newly reveals becomes low-confidence until reacquired. A moved
occluder refreshes the light-visibility field and nearby shadow boundaries. A
changed light invalidates direct and indirect radiance subslots for the surfaces
inside its influence region, but leaves geometry, normals, albedo, and much of
the BVH intact. A material edit invalidates BRDF-dependent radiance while
leaving visibility and geometry reusable.

\begin{table*}[t]
\centering
\footnotesize
\caption{Cache protocol for dynamic scene changes. The default action is
partial invalidation and sparse reacquisition, not whole-world rebuild.}
\label{tab:cache-invalidation-protocol}
\begin{tabular}{p{0.16\textwidth}p{0.26\textwidth}p{0.22\textwidth}p{0.24\textwidth}}
\toprule
\textbf{Event} & \textbf{Invalidate} & \textbf{Reuse} & \textbf{Refresh kernel or oracle} \\
\midrule
Camera motion & screen projection, disoccluded pixels, history confidence & static geometry, material, light epochs & TAA reprojection, depth/ID rejection, sparse fill for holes \\
Rigid object motion & object TLAS transform, camera occlusion, object shadow footprint & object-local shape/material Gaussians & transform Gaussians, update TLAS, trace camera/shadow seeds in dirty tiles \\
Dynamic occluder & light visibility and shadow edges & surface material, static indirect priors outside influence & shadow-map/ray refresh, confidence decay behind moving occluder \\
Light pose/intensity/color change & direct-light subslots, affected indirect radiance, light epoch & geometry, normals, albedo, BVH, visibility where unchanged & per-light relighting if transport exists; otherwise sparse path tracing \\
Material/texture edit & material epoch, BRDF-dependent direct and indirect radiance & geometry, object pose, camera visibility & recompute shading and retrace affected indirect paths \\
Topology/deformation & local geometry epoch, covariance, normals, visibility & unrelated static world regions & local Gaussian rebuild or simulator oracle samples \\
\bottomrule
\end{tabular}
\end{table*}

This protocol naturally produces an online update loop:
\begin{enumerate}
  \item advance physics and scene state, producing $\Delta_t$;
  \item update dynamic transforms and local Gaussian frames;
  \item construct camera, shadow, and radiance dirty masks;
  \item decay confidence for invalidated Gaussian/radiance entries;
  \item acquire sparse samples only in dirty masks or high-variance regions;
  \item update Gaussian geometry, visibility, and radiance subslots;
  \item reconstruct the final sensor output with uncertainty gates.
\end{enumerate}
The expensive static world is rebuilt only when the geometry or material change
is so large that local updates no longer cover the error. This is the
Physical-AI version of temporal rendering: most frames are not new worlds, but
new measurements of a partially changed world.

\subsection{Throughput-bound, latency-tolerant}

A real-time renderer must finish a frame in $\sim$16~ms or it fails.
A SDG pipeline must produce a target number of labeled images per dollar
or per joule; per-image latency is bounded only by the human iteration
loop and downstream training pipeline. If a software renderer takes
50~ms per image but uses one-tenth the silicon cost of an RTX-class
solution, it wins on $\$\slash\textrm{image}$ even though it loses on
frame time.

Table~\ref{tab:budgets} summarizes the gap between real-time and SDG
per-image budgets. Real-time renderers operate in single-digit-to-tens
of milliseconds, with the constraint that exceeding the budget is
visible to a human user as a dropped frame. SDG pipelines are sized by
the dataset they need to deliver: a typical robotics training run
consumes between $10^5$ and $10^7$ labeled images, the renderer is
allowed minutes per image if necessary, and the binding constraint is
total wall-clock time, dollars, or joules to produce the dataset
rather than per-image latency.

\begin{table}[t]
\centering
\caption{Per-image budgets, real-time vs.\ SDG. Real-time numbers are
the minimum frame budgets needed to avoid visible drops; SDG numbers
are the upper bounds at which a typical training run remains
practical. The relevant figure of merit shifts from \emph{latency} to
\emph{aggregate throughput per dollar / per joule}.}
\label{tab:budgets}
\footnotesize
\setlength{\tabcolsep}{2.5pt}
\renewcommand{\arraystretch}{1.12}
\begin{tabular}{@{}>{\raggedright\arraybackslash}p{0.29\columnwidth}
                >{\raggedright\arraybackslash}p{0.34\columnwidth}
                >{\raggedright\arraybackslash}p{0.27\columnwidth}@{}}
\toprule
& \textbf{Real-time} & \textbf{SDG} \\
\midrule
Per-image budget         & $\leq 16.7$ ms (60\,Hz) & $10^2$--$10^4$ ms \\
                         & $\leq 11.1$ ms (90\,Hz VR)  & \\
Tolerable variance       & low (jitter visible)        & high \\
Resolution               & 1080p--4K display-bound     & 256$^2$--1024$^2$ \\
Images per run           & $\sim$1 (continuous)        & $10^5$--$10^7$ \\
Dominant constraint      & per-frame latency           & throughput, \$/img \\
Hardware preference      & raster + RT cores           & available FLOPS \\
\bottomrule
\end{tabular}
\end{table}

A renderer that produces a $1024{\times}1024$ image in 50~ms is
unusable for real-time, but on a single H100 it would deliver roughly
$1.7{\times}10^6$ images per day. At the same throughput, a 16-GPU pod
produces a 27~M-image dataset per day, which exceeds the size of every
publicly released robotics training dataset to date.

\subsection{Quality requirements}

SDG renderers do not need to convince a human that the image is real;
they need to teach a model that the image is informative. Modern
research~\cite{tobin2017domain,nvidia2024cosmos} suggests that
\emph{diversity}, \emph{labeling fidelity}, and \emph{controlled
randomization} matter more than the marginal photorealism that justifies
RT-core hardware. A 3DGS rendering of a real scene captured with a
phone, augmented with diverse synthetic objects, can be more useful for
training than a mathematically prettier path-traced render of an
unrealistic asset library.

\subsection{Hardware availability and cost}

Table~\ref{tab:hardware} compares the compute economics of the
graphics-class silicon SDG normally targets against the AI-accelerator
silicon teams already own. Two effects compound. First, the
graphics-class lineup (L40, RTX 6000 Ada) ships in much smaller
volumes than the data-center training accelerators: NVIDIA's H100
shipments alone exceeded an estimated three to four million units in
2024, while the L40\slash L40S\slash RTX 6000 Ada production volume is
smaller by roughly an order of magnitude. Second, on a per-unit basis
H100 and B200 deliver $4$--$10{\times}$ more raw FLOPS per dollar of
spot-market rental than graphics-class
parts~\cite{nvidia-hopper-h100,nvidia-blackwell-b200}.

\begin{table}[t]
\centering
\caption{Illustrative per-device specifications and rough order-of-magnitude
spot-market prices~\cite{nvidia-hopper-h100,nvidia-blackwell-b200,nvidia-l40,google-tpu-v5}.
TFLOPS columns report dense BF16\slash FP16 throughput (the regime SDG
kernels run in), not sparse or FP8. Spot prices are illustrative and
fluctuate; we report them only to anchor the cross-class ratios.}
\label{tab:hardware}
\small
\begin{tabular}{@{}lcccc@{}}
\toprule
& \textbf{TFLOPS} & \textbf{HBM/Mem} & \textbf{TDP} & \textbf{Spot \$/hr} \\
& \textit{(BF16)} & (GB) & (W) & \textit{(rough)} \\
\midrule
\textit{AI-accelerator class} & & & & \\
NVIDIA H100 SXM   &  $\sim$990  & 80   & 700  & 2--4  \\
NVIDIA B200       & $\sim$2{,}500 & 192 & 1000 & 4--8  \\
Google TPU v5e    &  $\sim$200  & 16   & 170  & 1--2  \\
AWS Trainium2     &  $\sim$1{,}300 & 96 & ---  & 1--3  \\
\midrule
\textit{Graphics class}       & & & & \\
NVIDIA L40\,/\,L40S         &  $\sim$360  & 48   & 300  & 1--2 \\
NVIDIA RTX 6000 Ada         &  $\sim$364  & 48   & 300  & 1--2 \\
\bottomrule
\end{tabular}
\end{table}

The qualitative point survives any uncertainty in the absolute numbers.
A team that wants to do SDG today is either provisioning a separate
fleet of graphics-class GPUs (competing with rendering studios for a
constrained supply), or running on consumer hardware (RTX 4090, 5090)
that is unstable in data-center deployments. AI compute, by contrast,
is the silicon those teams already operate at scale for training. If
the renderer can run there, the marginal cost of the SDG run is the
opportunity cost of pre-existing capacity rather than the capex of a
new fleet.

\subsection{Economic compatibility with AI data centers}
\label{sec:economic-case}

The strongest economic case for vkGSplat is not merely that an H100 has high
FLOPS. It is that the AI data center is the platform on which organizations are
already spending, staffing, monitoring, and optimizing. Large AI operators now
commit tens of billions of dollars per year to AI-capable infrastructure:
Microsoft stated that it was on track to invest approximately \$80B in FY2025 to
build AI-enabled datacenters; Alphabet expected approximately \$75B in 2025
capital expenditures; and Meta's late-2025 outlook put capital expenditures at
\$70--72B, with infrastructure capacity identified as the central requirement
for its AI opportunities~\cite{microsoft-ai-datacenter-2025,alphabet-q4-2024-capex,meta-q3-2025-capex}.
Amazon's 2024 shareholder letter makes the same structural point from another
angle: AI demand requires datacenters, chips, and hardware to be bought upfront,
and the firm is investing in custom AI silicon such as
Trainium~\cite{amazon-shareholder-letter-2024}.
Omdia's Data Center World 2026 outlook adds a second, more physical constraint:
AI data centers are being reorganized around rack-scale systems, 800VDC power
distribution, battery energy storage, self-generation, liquid cooling, and
MW-class cooling distribution units~\cite{omdia-dcw-oas-2026}. This matters for
vkGSplat because the cost of a separate SDG graphics island is no longer only
the purchase price of GPUs. It is also power architecture, cooling qualification,
space, operational tooling, and the opportunity cost of not using the
AI-accelerator racks whose electrical and thermal infrastructure is already
being designed.

This is a compatibility argument, not a claim that CUDA software rendering must
beat a mature RTX stack at every graphics microbenchmark. NVIDIA's current SDG
approach is coherent: use RTX hardware for acquisition, Omniverse/Isaac for
simulation and scene tooling, OVX for graphics-heavy digital-twin and
simulation workloads, DGX/DGX Cloud for AI-factory training and inference, and
AI reconstruction or world models where they fit~\cite{nvidia-ovx,nvidia-dgx-cloud}.
vkGSplat targets the complementary economic case. If the organization is
already committing its capital, datacenter power, procurement, cluster
schedulers, storage paths, observability, and software qualification around
AIDCs, then creating a separate graphics island can be more disruptive than
accepting a less specialized but colocated compute implementation.

This changes the SDG question. A graphics-specific SDG stack asks an
organization to buy, operate, and schedule a second production substrate:
graphics GPUs, graphics drivers, graphics observability, graphics queue
management, renderer-specific deployment workflows, and a data path that moves
rendered examples back into AI training storage. That may be correct for
high-end visual effects or game-content pipelines, where graphics hardware is
the primary factory. It is much harder to justify inside a robotics or foundation
model organization whose dominant capital plan is already AI compute. In such
an organization, the expensive thing is not only the GPU; it is disrupting the
AIDC buildout with another procurement, deployment, and operations stack.

The economic tension is therefore not ``NVIDIA versus non-NVIDIA.'' It is
whether the SDG factory should be a heterogeneous two-cluster ecosystem or a
unified software-defined workload on the AI data center. In the heterogeneous
model, an OVX/RTX/Omniverse cluster produces simulated physical evidence and a
DGX-class AI cluster consumes it. In the unified model, the Vulkan-compatible
acquisition contract, seed rendering, reconstruction, sensor modeling, world
models, and training all share the same scheduler, accelerator pool, tensor
runtime, and storage locality. vkGSplat is the second model: it tries to make
graphics and physical-data acquisition programmable enough that the massive
compute resources already installed for AI can absorb SDG without a separate
graphics-specialized island.

The compute-first Vulkan path reframes SDG as an AI data-center application.
The workload becomes a batch or streaming job with familiar properties:
containerized kernels, CUDA/PyTorch tensors, object-store input and output,
standard cluster scheduling, standard quota and billing, model-side telemetry,
and the same accelerator fleet used for training, inference, and world-model
rollouts. This is economically important even if a graphics accelerator wins
some individual visibility tests. A specialized renderer can lose at the system
level if it lowers per-ray cost but adds a separate fleet, lower utilization,
extra data movement, separate software qualification, and a procurement cycle
that cannot track AI-infrastructure scale. The success criterion is therefore
not ``out-render RTX.'' It is ``make SDG fit the AIDC operating model while
remaining good enough on physical fidelity, throughput, labels, and cost.''

\begin{table}[t]
\centering
\small
\caption{Economic interpretation of the vkGSplat strategy. The claim is not
that fixed-function graphics is obsolete or that RTX SDG is the wrong baseline;
it is that SDG becomes easier to scale when it does not disrupt the AI
data-center buildout.}
\label{tab:economic-case}
\begin{tabular}{p{0.29\textwidth}p{0.30\textwidth}p{0.31\textwidth}}
\toprule
\textbf{Decision axis} & \textbf{Heterogeneous OVX/graphics + DGX AI clusters} & \textbf{Unified AI-data-center SDG workload} \\
\midrule
Capital plan & Dedicated graphics procurement and refresh cycle & Rides existing AI accelerator capex and depreciation schedule \\
Utilization & Must keep a rendering fleet busy enough to justify ownership & Can fill training gaps, off-peak windows, and batched data-generation queues \\
Software stack & Renderer-specific deployment, driver, profiling, and observability path & CUDA/PyTorch containers, standard cluster telemetry, common job scheduler \\
Graphics hardware assumption & Relies on RTX-class fixed-function graphics where acquisition cost, driver queues, and framebuffer contracts are natural & Uses software rendering to preserve the Vulkan acquisition contract while running on AI accelerators without graphics-specific silicon \\
Reconstruction control & DLSS-class reconstruction is powerful but proprietary, product-tuned, and game/display-oriented & Uses integrated 3DGS state as an open DLSS-class counterpart: Gaussian/radiance history, temporal gates, SR, denoising, frame interpolation, and optional learned residuals \\
Data movement & Simulation cluster emits data that must be moved to AI training storage & Acquisition, reconstruction, world models, and training can share device or nearby storage paths \\
Optimization target & Per-frame or per-ray graphics performance & Physical evidence, labels, sensor streams, and task accuracy per dollar or joule \\
Organizational fit & Treated as a specialized simulation/graphics service adjacent to the AI factory & Treated as another AI workload beside training, inference, evaluation, and data processing \\
\bottomrule
\end{tabular}
\end{table}

\begin{table}[t]
\centering
\small
\caption{External AIDC pressures and their architectural interpretation for
vkGSplat. The Omdia figures are used as market evidence, not as hard physical
laws; the design consequence is that SDG should minimize disruption to the
AI-data-center power, cooling, and scheduling envelope.}
\label{tab:omdia-aidc-pressures}
\begin{tabular}{p{0.30\textwidth}p{0.29\textwidth}p{0.31\textwidth}}
\toprule
\textbf{AIDC pressure} & \textbf{Omdia 2026 signal} & \textbf{vkGSplat implication} \\
\midrule
IT and AIDC spending scale & \$6.07T total IT spending in 2026; data-center investment scenario approaching \$1.9T by 2030 & SDG should be packaged as an ordinary AI-data-center workload, not a bespoke rendering island \\
Budget growth and adoption pressure & 72.9\% of organizations expect IT-budget increases in 2026; AI usage and model demand remain strong & Synthetic data generation must share the same scheduling, quota, telemetry, and utilization logic as training and inference \\
Power and cooling become first-order & Rack-scale AI systems, 800VDC, BESS, self-generation, and liquid cooling move into the data-center roadmap & Adding graphics-specific hardware has physical-infrastructure cost, not just silicon cost; colocated compute paths become strategically valuable \\
Physical AI token/data demand & Omdia positions Physical AI as a later wave that may demand even more tokens than current generative and agentic workloads & SDG must be amortized, explicit-state, and reconstruction-heavy; it cannot rely on full-fidelity acquisition for every sample \\
\bottomrule
\end{tabular}
\end{table}

This argument also explains why vkGSplat keeps Vulkan as the front-end contract.
The goal is not to discard graphics semantics; it is to make those semantics
deployable where the economic gravity already sits. A Vulkan-compatible
acquisition contract lets existing assets and rendering abstractions survive,
while the backend can be lowered to CUDA kernels, tensorized reconstruction,
or eventually other AI-accelerator runtimes. If a site has abundant RTX or L40
hardware, hardware Vulkan remains the natural acquisition backend. If the site
has a large H100/B200/TPU/Trainium-style fleet and a hard time creating a new
graphics island, vkGSplat makes SDG a software workload on the AI platform.

A back-of-envelope energy budget reinforces the argument. At a 50-ms
target image budget on an 80\% utilized H100 (700~W TDP), the
rendering energy cost per image is roughly $0.3$~J. At
$10^7$ images, that is $\sim 0.8$ kWh of rendering work --- a
rounding error against the multi-megawatt-hour cost of training the
downstream model.

The same calculation should not be read as a promise that software acquisition
is always more energy efficient than fixed-function graphics. It is a system
placement argument. If the site already has a liquid-cooled, power-qualified AI
cluster with idle or batch-fill capacity, the incremental SDG job may be
economically preferable even when each visibility query is less specialized. If
the site already has an RTX/L40/Omniverse production substrate, hardware
graphics remains the correct baseline. vkGSplat is valuable precisely because it
allows both paths to share a Vulkan seed-buffer contract and a downstream
Gaussian/PyTorch reconstruction stack.

% =====================================================================
\section{System design}\label{sec:design}
% =====================================================================

vkGSplat's software-CUDA branch is structured as three layers
(Fig.~\ref{fig:arch}). In the more general acquisition-neutral architecture, a
hardware Vulkan renderer is a sibling branch: it may run the normal graphics
pipeline and export the same seed-buffer ABI to the PyTorch/vkGSplat
reconstruction stack. The figure therefore describes the distinctive vkGSplat
implementation path, not the only valid acquisition path.

\begin{figure}[t]
\centering
\begin{tikzpicture}[
    font=\footnotesize,
    box/.style    ={draw, rounded corners=2pt, align=center,
                    minimum width=6.4cm, minimum height=0.85cm,
                    inner sep=2pt},
    sub/.style    ={draw, rounded corners=1pt, align=center,
                    minimum width=1.85cm, minimum height=0.65cm,
                    inner sep=1pt, font=\scriptsize},
    arrow/.style  ={-Latex, thick},
    layerlabel/.style={font=\scriptsize\itshape, anchor=west}
]
% --- top: applications ---
\node[box, fill=black!5] (apps) {Vulkan applications\\ \scriptsize Isaac Sim, BlenderProc, custom robotics SDG harnesses};

% --- layer 1: front-end ---
\node[box, fill=blue!8, below=0.55cm of apps] (vk) {%
    \textbf{Vulkan front-end (ICD)}\\
    \scriptsize subset of Vulkan 1.3 + RT + external memory
};

% --- layer 2: IR ---
\node[box, fill=green!8, below=0.55cm of vk] (ir) {%
    \textbf{Compute IR}\\
    \scriptsize SPIR-V $\to$ LLVM/MLIR; command buffer record \& replay
};

% --- layer 3: backends, three side by side ---
\node[sub, fill=orange!12, below=0.55cm of ir, xshift=-2.1cm] (cuda) {CUDA\\\scriptsize H100,B200};
\node[sub, fill=orange!12, right=0.18cm of cuda] (triton) {Triton\,/\,SYCL\\\scriptsize MI300, PVC};
\node[sub, fill=orange!12, right=0.18cm of triton] (ainative) {AI-native\\\scriptsize TPU, Trainium};

% --- bracket label for backends ---
\node[layerlabel, left=0.1cm of cuda, anchor=east, align=right]
    {\textbf{Compute}\\\textbf{backends}};

% --- kernel modules below backends ---
\node[box, fill=red!6, below=0.55cm of triton, minimum width=6.4cm] (kernels)
    {\textbf{Rendering kernels}\\%
     \scriptsize gsplat-style 3DGS~|~nvdiffrast-style raster seed~|~BVH ray traversal~|~shadow / denoise};

% --- silicon ---
\node[box, fill=gray!8, below=0.55cm of kernels] (hw)
    {\textbf{AI-accelerator silicon}\\\scriptsize tensor\,/\,SIMT compute, HBM, NVLink};

% --- arrows down ---
\draw[arrow] (apps) -- node[right]{\scriptsize Vulkan API} (vk);
\draw[arrow] (vk)   -- node[right]{\scriptsize SPIR-V + cmd buffers} (ir);
\draw[arrow] (ir.south) -- ++(0,-0.18) -| (cuda.north);
\draw[arrow] (ir.south) -- ++(0,-0.18) -| (triton.north);
\draw[arrow] (ir.south) -- ++(0,-0.18) -| (ainative.north);
\draw[arrow] (cuda.south)    |- ($(kernels.north)+(0,0.12)$) -- (kernels.north);
\draw[arrow] (triton.south)  -- (kernels.north);
\draw[arrow] (ainative.south)|- ($(kernels.north)+(0,0.12)$) -- (kernels.north);
\draw[arrow] (kernels) -- (hw);

% --- side annotation: existing graphics path crossed out ---
\node[draw, dashed, rounded corners=2pt, align=center, font=\scriptsize,
      right=0.45cm of vk, fill=gray!5, minimum height=2.7cm, text width=2.0cm]
      (notused)
      {\emph{Bypassed:}\\rasterizer\\ROPs\\RT cores};
\draw[dashed,gray] (vk.east) -- (notused.west);

\end{tikzpicture}
\caption{vkGSplat architecture. Applications retain the standard Vulkan
API; the implementation lowers SPIR-V and command buffers to a
backend-agnostic compute IR (LLVM\,/\,MLIR), which is dispatched to
CUDA, Triton\,/\,SYCL, or AI-accelerator-native backends. The
rendering kernels --- 3DGS rasterizer, software tile rasterizer,
BVH ray traversal, denoising --- run as compute on tensor- and
SIMT-rich silicon. The fixed-function graphics path is bypassed in this
software backend, while a hardware Vulkan acquisition branch can still emit the
same seed-buffer contract.}
\label{fig:arch}
\end{figure}

\begin{enumerate}
    \item \textbf{Vulkan front-end.} A Vulkan loadable ICD that
    implements a \emph{conformance-targeted subset} of Vulkan~1.3 plus
    selected ray-tracing and external-memory extensions. Mesh shaders
    are treated as the preferred programmable primitive-expansion path
    for Gaussian splats. Tessellation and geometry shaders are not in
    the v1 target: their useful role in this system, expanding compact
    primitives into screen-space work, is subsumed by mesh shaders and
    then by the tile-based software backend. Applications link the
    standard Vulkan loader and need no modification.

    \item \textbf{Intermediate representation (IR).} SPIR-V shaders and
    Vulkan command buffers are lowered into a backend-agnostic compute
    IR. We propose to base this IR on existing infrastructure ---
    LLVM\slash MLIR for shader code, a record-and-replay command-buffer
    IR for pipeline state --- so that the system rides on existing
    compiler ecosystems rather than inventing one.

    \item \textbf{Compute backend.} The IR is lowered to one of several
    backends: CUDA (reference), Triton or SYCL for portability, and
    eventually AI-accelerator-native paths (XLA HLO for TPU, NeuronCore
    for Trainium, etc.). The backend implements the rendering kernels:
    rasterization, ray traversal, shadow mapping, and
    Gaussian-splat blending.
\end{enumerate}

\subsection{Concrete CUDA software stack}

The concrete software stack is now deliberately organized around
CUDA-resident \emph{seed buffers}.  Existing Vulkan programs, beginning
with Wicked Engine workloads, remain the application input.  A generic
vkGSplat-compatible pipeline may obtain those seed buffers from a hardware Vulkan
renderer through interop or capture; the distinctive vkGSplat backend captures
the command-buffer, descriptor, SPIR-V, camera, and geometry state, then routes
draw and ray-dispatch work into a software CUDA renderer rather than a
fixed-function graphics queue.  The output of either renderer is not merely a
display image: it is a compact G-buffer-like seed containing color, depth,
primitive identity,
barycentrics or local coordinates, motion-reconstruction inputs, and
optional sample statistics.  Those buffers remain on the CUDA device and
become the direct input to downstream CUDA kernels for screen-space
motion, history reprojection, denoising, temporal antialiasing, and
eventually Gaussian reconstruction.

Three existing CUDA renderers define the source-level reference points.
For triangle visibility, vkGSplat follows the
\textsc{nvdiffrast}/CudaRaster split~\cite{laine2020nvdiffrast}: triangle
setup, bin rasterization, coarse rasterization, and fine rasterization
are separate kernels that produce per-pixel visibility facts such as
depth, triangle identity, and barycentric coordinates.  For canonical
Gaussian visibility, vkGSplat follows the GraphDECO CUDA rasterizer
released with 3DGS~\cite{kerbl20233dgs}: frustum culling and
preprocessing, duplicate Gaussians into tile/depth-keyed work items,
identify tile ranges, and launch tile-local \texttt{renderCUDA}
compositing.  For a broader maintained library reference, vkGSplat also
tracks \textsc{gsplat}~\cite{ye2024gsplat}: fused EWA projection, packed
and unpacked tile intersections, tile-offset construction, flattened
Gaussian IDs, and tile-local front-to-back alpha compositing.  The
important design point is that these paths expose the same kind of
contract --- tile-indexed device buffers --- so the reconstruction stage
does not care whether the seed came from triangle rasterization,
ray-traced visibility, or Gaussian splatting.

This turns the CUDA backend into a producer--consumer pipeline:
\[
\begin{gathered}
\text{Vulkan/Wicked stream}\\
\downarrow\\
\text{CUDA software seed renderer}\\
\downarrow\\
\text{CUDA reconstruction kernels}\\
\downarrow\\
\text{optional CUDA 3DGS renderer}
\end{gathered}
\]
Debug viewers may copy or interop-present the final image, but the
production SDG path should avoid host readback between these stages.
The same buffer ABI is also what CPU and Metal backends must emulate for
correctness testing.

\subsection{PyTorch as orchestration hub and world-model interface}
\label{sec:pytorch-world-hub}

The practical control plane should be PyTorch, while the performance-critical
inner loops remain CUDA, Vulkan-facing command lowering, and explicit
vkGSplat kernels. This is not a proposal to express rasterization, BVH
traversal, or Gaussian tile compositing as ordinary eager tensor algebra. Many
of those kernels need scatter, gather, atomics, prefix sums, sorting, queues,
and visibility-dependent branches. PyTorch is instead the hub that owns
experiment scheduling, tensor lifetimes, autograd where useful, distributed
training, data loading, and model inference. CUDA extensions, nvdiffrast-style
operators, Warp-style simulation kernels, and external-memory Vulkan interop
then exchange CUDA-resident tensors without a host copy
\cite{paszke2019pytorch,nvidia-warp}.

This also gives a clean place to connect physical simulation, physics-based
rendering, sensor modeling, vkGSplat reconstruction, and learned world models.
The physics engine advances state; the renderer and sensor models acquire
sparse but physically grounded evidence; vkGSplat fuses that evidence into a
persistent geometric/radiance state; optional neural modules denoise,
upscale, interpolate, or predict futures. PyTorch coordinates the graph and
training losses, but the authority of each value still comes from its source:
calibrated sensors and path tracing for measurements, simulation for dynamics,
Gaussians for amortized reconstruction, and world models for learned priors or
counterfactual proposals.

\begin{figure*}[t]
\centering
\begin{tikzpicture}[
    node distance=0.68cm and 0.72cm,
    every node/.style={font=\footnotesize},
    box/.style={draw, rounded corners=2pt, align=center, minimum width=2.65cm, minimum height=0.68cm},
    wide/.style={draw, rounded corners=2pt, align=center, minimum width=3.35cm, minimum height=0.68cm},
    arrow/.style={-Latex, thick}
]
\node[box] (phys) {Physics engine\\state, contacts, materials};
\node[box, right=of phys] (sensor) {Path tracer and sensors\\sparse physical evidence};
\node[wide, right=of sensor] (torch) {PyTorch hub\\tensors, streams, training, policy};
\node[box, right=of torch] (vk) {vkGSplat CUDA\\Gaussian state and resolver};
\node[box, below=of torch] (wm) {World model\\prior, prediction, counterfactuals};
\node[box, right=of vk] (out) {SDG outputs\\images, labels, trajectories};

\draw[arrow] (phys) -- (sensor);
\draw[arrow] (sensor) -- (torch);
\draw[arrow] (torch) -- (vk);
\draw[arrow] (vk) -- (out);
\draw[arrow] (torch) -- (wm);
\draw[arrow] (wm) -- (torch);
\draw[arrow] (vk.south) |- (wm.east);
\draw[arrow] (torch.north) to[bend right=18] node[above] {adaptive oracle requests} (sensor.north);
\end{tikzpicture}
\caption{PyTorch-centered orchestration keeps the programmable AI stack in
charge of tensors, training, and scheduling, while vkGSplat remains the
high-throughput CUDA reconstruction backend. World models are compatible with
this design when they act as priors, rollout generators, or proposal
mechanisms, and when uncertainty can trigger additional physical acquisition.}
\label{fig:pytorch-world-hub}
\end{figure*}

\begin{table*}[t]
\centering
\footnotesize
\caption{Compatible roles in a PyTorch-centered vkGSplat stack.}
\label{tab:pytorch-world-roles}
\begin{tabular}{p{0.16\textwidth}p{0.36\textwidth}p{0.38\textwidth}}
\toprule
Component & Primary contract & Failure if misused \\
\midrule
Physics and sensors & Authoritative dynamics, contacts, calibration, and sparse physical measurements & Treating reconstruction artifacts as physical truth \\
Path tracer & Low-sample but physically grounded radiance acquisition with adequate bounce depth & Cutting path depth until missing light transport is unrecoverable \\
vkGSplat state & Persistent 3D/4D cache for geometry, radiance, variance, visibility, and motion & Hallucinating unsupported Gaussians without confidence gates \\
World model & Learned prior, counterfactual rollout, scene proposal, and missing-context predictor & Replacing calibrated physics where safety-critical accuracy is required \\
PyTorch hub & Tensor ownership, scheduling, training, neural inference, and interop & Forcing irregular visibility kernels into inefficient dense tensor form \\
\bottomrule
\end{tabular}
\end{table*}

World-model research is therefore compatible with vkGSplat, not a competitor
to it. Classical world models learn compact latent dynamics for control
\cite{ha2018}; Genie-style and Cosmos-style systems scale that idea toward
interactive or generative environments for embodied agents
\cite{bruce2024,deepmind2024genie2,nvidia2024cosmos}. The word ``replace''
needs care here. A learned world model can replace many expensive simulator
queries when the goal is exploration, data diversification, counterfactual
rollout, photoreal transformation, or policy pretraining under a measured
validation envelope. It should not be treated as replacing the calibrated
physics engine, renderer, or sensor model where the required output is a
measurement with known error bars. NVIDIA's own Cosmos positioning reflects
this split: Cosmos Transfer is described as pairing with simulation frameworks
such as CARLA and Isaac Sim, while Cosmos Predict generates controllable
predictive worlds for downstream tasks~\cite{nvidia2024cosmos}. Genie-style
models similarly emphasize playable action-controllable environments for agent
research, not certified contact, optics, or sensor calibration.

Those systems can propose trajectories, synthesize environment variation,
provide learned temporal priors, or generate counterfactual rollouts for policy
training. vkGSplat gives that learned stack an explicit 3D/4D state and a
physically grounded acquisition interface. The most useful architecture is a
closed loop: the world model predicts or proposes; vkGSplat projects and
reconstructs explicit visual state; calibrated simulation and sensor/path-tracing
oracles refresh regions where uncertainty, rare events, or task risk demand
physical evidence.

\subsection{Vulkan front-end}

We target a deliberate \emph{subset}: graphics pipelines, compute
pipelines, command buffers, descriptor sets, mesh shaders, indirect
dispatch/draw for GPU-driven splat counts, swapchain (optional, for
debug viewer), timeline semaphores, and external memory. We initially
omit tessellation shaders, geometry shaders, sparse binding, and
presentation-engine optimization. SDG does not need them, and for
Gaussian splatting the tessellation/geometry stages are a worse fit
than mesh shaders followed by tile-local software shading.

\subsection{Lowering rasterization}

We lower the classical raster pipeline to a tile-based software
rasterizer in the spirit of Laine and
Karras~\cite{laine2011software}, with each output tile assigned to a
threadblock\slash workgroup that performs binning, depth test, and
fragment shading.  The immediate implementation reference is
\textsc{nvdiffrast}'s CudaRaster path~\cite{laine2020nvdiffrast}, whose
CUDA stages are explicitly separated into triangle setup, bin raster,
coarse raster, and fine raster kernels.  vkGSplat adopts that staging
but changes the product: instead of a differentiable PyTorch image
operator, the rasterizer emits CUDA-resident seed buffers for SDG
reconstruction. The same organization is used for the 3DGS fast path:
Gaussians are projected once, compacted into screen-space tile lists,
and then blended by tile-local workers. This is the crucial portability
choice. A tile maps naturally to a CUDA cooperative thread array, a SYCL
workgroup, or a TPU/Trainium-style block of dense compute; it also keeps
memory traffic local enough that the backend can be retargeted without
recreating a fixed-function raster pipeline in disguise.

\subsection{Lowering ray tracing and shadow mapping}

Ray-tracing pipelines (\texttt{VK\_KHR\_ray\_tracing\_pipeline}) lower to
software BVH traversal kernels following the
Aila--Laine~\cite{aila2009understanding} pattern, optionally calling
into Embree-like tree construction routines on the host. Shadow maps
become depth-only raster passes followed by a comparison kernel; for
high-quality shadows we expect ray-traced shadows to be cheaper than
classical shadow-map filtering at SDG quality targets.

The seed path is intentionally low-resolution and low-sample, but not
intentionally shallow. Each frame exports color, depth, primitive identity,
camera state, sample count, path-depth statistics, and variance estimates into
the temporal reconstruction contract. Adjacent frames then provide the
screen-space motion map needed for history lookup; depth and primitive identity
reject disocclusions and visibility changes; a small SVGF-style filter removes
Monte Carlo noise before any optional learned reconstruction stage is
considered. The acquisition policy should first reduce samples per pixel and
seed resolution, then spend extra samples adaptively where variance, missing
path-depth support, or sensor confidence says the Gaussian history is weak.

\subsection{Paused optional 3DGS fast path}\label{sec:v1}

The project previously exposed a non-standard Vulkan extension
(\texttt{VK\_VKGSPLAT\_gaussian\_splatting}) that lets applications submit
Gaussian primitives directly. Internally, this dispatches to a
3DGS rasterizer akin to the GraphDECO CUDA rasterizer released with
3DGS~\cite{kerbl20233dgs}, but extended for offline SDG (deterministic
ordering, deeper sample budgets per pixel, optional denoising).  The
CUDA kernel organization is cross-checked against
\textsc{gsplat}~\cite{ye2024gsplat}: fused EWA projection, tile
intersection, tile-offset construction, flattened Gaussian IDs, and
tile-local pixel compositing. The Vulkan-facing primitive expansion uses mesh shaders:
one workgroup processes a batch of Gaussians, computes view-dependent
color and projected covariance once per splat, and emits the small
screen-space proxy needed by the fragment/tile stage. The production
software backend then keeps the same logical structure but executes it
as explicit tile-binned compute: preprocess, cull, assign splats to
tiles, sort or approximately order within each tile, and blend pixels.

This is why vkGSplat drops tessellation and geometry shaders from the
initial design. Tessellation expands surfaces; geometry shaders expand
or amplify primitives late in the traditional graphics pipeline. For
Gaussian splats, both jobs are better represented by mesh shaders at the
API boundary and by tile-local software work in the backend. A system
that ships this path on day one is immediately useful for captured real
environments and for Lyra-style generated Gaussian scenes. For the
current milestone sequence, however, this path is paused behind a build
flag so the default implementation can focus on Vulkan ray-tracing seed
frames and CUDA lowering.

% =====================================================================
\section{Backends}\label{sec:backends}
% =====================================================================

\textbf{CUDA.} The reference implementation, targeting NVIDIA H100 and
B200. CUDA is mature, has a rich kernel ecosystem (cuBLAS, CUB, Thrust,
OptiX-as-library), and provides external-memory and external-semaphore
interop with Vulkan~\cite{nvidia2020vkcuda}.

\textbf{Triton\slash SYCL.} A second backend in Triton or SYCL would
demonstrate portability and allow deployment on AMD MI300, Intel Ponte
Vecchio, and any future accelerator with an LLVM-class compiler.

\textbf{Metal.} Apple Silicon provides a useful local portability
target. A native Metal backend is not the deployment target, but it
tests whether the reconstruction contracts are genuinely compute
portable rather than CUDA-shaped by accident. We use it first for
temporal reconstruction and denoising on the M-series GPU, then compare
results against the CPU reference before porting the same contracts to
CUDA.

\textbf{AI-accelerator-native.} TPU (XLA HLO), Trainium (NeuronCore SDK),
Cerebras, and Groq each have native compilation paths. Lowering rendering
kernels to these stacks is non-trivial but, for \emph{compute-dominant}
kernels (3DGS rasterization, BVH traversal), feasible. The rendering
problem becomes structurally close to a sparse-attention problem, which
these chips already optimize for.

% =====================================================================
\section{Evaluation plan}
% =====================================================================

We will evaluate vkGSplat along three axes.

\textbf{Throughput.} Images per second at $512^2$ and $1024^2$
resolution on three scene families:
(i)~the standard novel-view-synthesis benchmarks --- Mip-NeRF~360,
Tanks~\& Temples, Deep Blending --- to permit comparison with
published 3DGS numbers~\cite{kerbl20233dgs};
(ii)~SDG-typical procedural scenes from ProcTHOR~\cite{deitke2022procthor}
and a representative subset of Isaac~Lab industrial-manipulation
environments, rendered with controlled domain randomization;
and (iii)~hybrid scenes consisting of a 3DGS-captured real environment
populated with synthetic foreground objects, which is the natural SDG
flow vkGSplat is built for.

\textbf{Cost \& energy.} Dollars-per-image and joules-per-image on a
range of accelerators (H100, B200, A100, L40, MI300, TPU~v5e). The L40
serves as a graphics-hardware baseline; the rest are pure compute. Because the
economic claim is about AIDC compatibility rather than peak ray throughput, we
also report queue fill, device utilization, host-device copies, cross-cluster
data movement, and whether acquisition, reconstruction, and training can share
one scheduler and storage locality.

\textbf{Downstream task accuracy.} Train a small object-detection or
manipulation policy on data generated by vkGSplat vs.\ data generated by
Isaac~Sim or BlenderProc, controlling for total training compute. The
hypothesis is that vkGSplat data is at least competitive on per-dollar
performance.

\textbf{Closed-loop data generation.} Run the same robot or perception model
through a static synthetic dataset, a random-domain-randomization dataset, and
an active vkGSplat loop where failures, high uncertainty, disocclusions, contact
events, and sensor-confidence failures request new acquisition. Report
downstream improvement per synthetic sample, failures recovered per GPU-hour,
oracle-refresh rate, uncertainty calibration, and the fraction of generated
data that is actually used by training. This is the direct test of
Eq.~\ref{eq:closed-loop-sdg-training}: SDG is valuable only if the next model
improves because the current model asked for better data.

Concretely, Table~\ref{tab:eval} fixes the matrix we will report.
Each cell is a (scene family, hardware, backend, renderer)
configuration; we report throughput, $\$\slash\textrm{image}$, and
$\textrm{J}\slash\textrm{image}$ as the primary metrics, with PSNR /
SSIM / LPIPS against held-out reference renders for image quality.
Downstream task accuracy is reported on a single curated robotics benchmark
(object-detection mAP on a held-out test set, plus a small manipulation-policy
success rate) rather than chasing a long tail of benchmarks; the goal is to
falsify the hypothesis ``models trained on vkGSplat data are competitive per
dollar with models trained on Isaac~Sim or BlenderProc data,'' not to
exhaustively characterize sim-to-real transfer. The closed-loop variant is
reported separately because it tests a stronger claim: the generator is no
longer a passive image factory but part of the training control loop.

We also evaluate the world-building claim directly. For each scene we separate
one-time build cost from per-frame acquisition cost:
\[
  C_{\mathrm{image}} =
  \frac{C_{\mathrm{build}}}{N_{\mathrm{episodes}} N_{\mathrm{frames}}}
  + C_{\mathrm{step}} + C_{\mathrm{acquire}} + C_{\mathrm{reconstruct}}.
\]
The experiment varies the reuse factor and the type of dynamic change. Static
scenes should show strong amortization from high-bounce precomputation and
Gaussian/radiance-cache warmup. Dynamic-occluder scenes should mostly increase
visibility and shadow refresh cost. Dynamic-light and material-edit scenes are
the hard case: color-only Gaussian caches should fail, while relightable
Gaussian state or sparse oracle retracing should preserve quality. We report
image metrics against high-spp references, but the pass/fail signal is
staleness: light leaks, wrong shadows, wrong indirect color, wrong sensor
statistics, and downstream task degradation after dynamic changes.
The ablation should compare (i) full rebuild every frame, (ii) static cache with
no invalidation, (iii) image-space invalidation only, and (iv) dependency-aware
Gaussian/radiance-cache invalidation with targeted oracle refresh. The expected
win condition is that option~(iv) approaches full-rebuild quality on dynamic
lighting and occlusion cases while retaining most of the amortization benefit of
the static cache.

\begin{table}[t]
\centering
\caption{Evaluation matrix. Each axis is independently varied; the
full Cartesian product is not run --- we ablate one axis at a time
against a reference configuration of \{H100, CUDA, 3DGS,
hybrid 3DGS+synthetic\}.}
\label{tab:eval}
\small
\begin{tabular}{@{}ll@{}}
\toprule
\textbf{Axis} & \textbf{Levels} \\
\midrule
Hardware  & H100, B200, A100, L40S, MI300, TPU v5e \\
Backend   & CUDA (ref), Triton, SYCL, XLA HLO (TPU) \\
Renderer  & 3DGS, software raster, software RT \\
Scene     & Mip-NeRF 360, T\&T, Deep Blending,\\
          & ProcTHOR, Isaac Lab, hybrid 3DGS+CAD \\
Resolution & $512^2$, $1024^2$ \\
\midrule
Metrics   & throughput (img/s), \$/img, J/img, \\
          & PSNR / SSIM / LPIPS, \\
          & downstream mAP, manipulation success \\
\bottomrule
\end{tabular}
\end{table}

% =====================================================================
\section{Discussion}
% =====================================================================

\textbf{When does this lose?} If the workload is dominated by
fixed-function operations (heavy alpha-tested foliage, complex
post-processing chains, high-fidelity displacement), software renderers
on compute will pay a large constant factor. Real-time interactive use
will stay on RTX hardware for the foreseeable future.

\textbf{Why not a new API?} A new graphics API for AI compute has been
proposed informally many times. The reason to keep Vulkan is ecosystem:
existing engines (Unreal's Vulkan RHI, Unity, Godot, Filament), tooling
(RenderDoc, NSight, ImGui), and middleware all speak Vulkan. Reusing
that surface means a vkGSplat backend benefits the ecosystem
asymmetrically.

\textbf{Strategic implications.} If a viable Vulkan-on-AI-compute path
exists, the SDG market is no longer constrained by the supply of
graphics silicon. This is structurally similar to what NVIDIA Omniverse
does for digital twins, but with the asset-distribution and
hardware-binding stories inverted: vkGSplat is open at the API, neutral
at the hardware, and assumes a software-defined data center as its
deployment target. It is also consistent with the direction the
graphics-hardware industry has been moving for half a decade
(Section~\ref{sec:dlss}): each successive generation of DLSS-class
neural reconstruction has shifted more of the rendering pipeline onto
tensor cores. vkGSplat is the architectural endpoint of that
trajectory --- a renderer in which \emph{all} of the work, seed and
reconstruction alike, runs on tensor-rich compute, and graphics
hardware is no longer in the loop.

\textbf{Limitations.} v1 will not be Vulkan-conformant. It will be
correct on a defined subset, on defined drivers, for defined SDG
workloads. Conformance is a long-term goal but not a pre-condition for
usefulness.

% =====================================================================
\section{Related work, expanded}
% =====================================================================

\textbf{Software rasterization.} Beyond the canonical Laine \&
Karras~\cite{laine2011software} result, FreePipe~\cite{liu2010freepipe}
demonstrated programmable parallel rasterization with multi-fragment
effects on GPUs, and Patney \& Owens~\cite{patney2010real} explored
adaptive subdivision on compute. \textsc{nvdiffrast}~\cite{laine2020nvdiffrast}
modernizes the CudaRaster lineage as a set of CUDA-accelerated
graphics primitives --- rasterization, interpolation, texturing, and
antialiasing --- and is the closest existing source-level reference for
vkGSplat's triangle seed renderer. These results predate the modern
tensor-core era but already establish that software-on-compute
rasterization is competitive on workloads where rasterization is not
the bottleneck --- the regime SDG sits in.

\textbf{Differentiable rendering.}
OpenDR~\cite{loper2014opendr} demonstrated that classical rasterizers
could be made approximately differentiable; Li et
al.~\cite{li2018edge} extended this to physically-based path tracing
with edge sampling. The same underlying compute substrate that powers
vkGSplat could host a differentiable variant in future work, allowing
SDG runs to feed gradients back into asset placement or material
parameters.

\textbf{Neural rendering surveys.} Tewari et al.~\cite{tewari2022advances}
surveys the broader space of neural rendering and clarifies that
3DGS sits within a continuum of explicit and implicit
representations. The defining property we exploit ---
compute-native rendering with no fixed-function dependencies --- is
shared with NeRF~\cite{mildenhall2020nerf} and Instant
NGP~\cite{mueller2022instant}, but 3DGS achieves rasterization speeds
that NeRF-class methods do not, which is why we choose it as the v1
renderer.

\textbf{3DGS extensions.} Mip-Splatting~\cite{yu2024mip}
addresses anti-aliasing across scale; 2DGS~\cite{huang20242dgs}
substitutes 2D primitives for tighter geometric fidelity; and
Spec-Gaussian~\cite{yang2024specgaussian} adds anisotropic specular
appearance. The original GraphDECO CUDA rasterizer released with
3DGS~\cite{kerbl20233dgs} is the canonical source reference for the
forward tile-compositing path, while \textsc{gsplat}~\cite{ye2024gsplat}
packages the CUDA projection, tiling, rasterization, batching, and
backward-pass machinery needed for large-scale Gaussian splatting.
vkGSplat targets the original 3DGS forward path first;
incorporating these variants is straightforward because all of them
remain compute-only.

\textbf{API translation precedents.} ANGLE~\cite{angle} translates
OpenGL ES to Direct3D, Metal, or Vulkan, and is the basis for
Chromium's WebGL implementation. VKD3D-Proton~\cite{vkd3dproton}
translates Direct3D~12 to Vulkan and is the engine that allows recent
Windows games to run on Linux under Proton. Both establish that a
shipped, correctness-targeted graphics-API translation layer is
feasible at production scale. vkGSplat's translation goes one step
further by changing not only the API but also the execution
substrate.

\textbf{CUDA--Vulkan interop.} NVIDIA's reference interop
samples~\cite{nvidia2020vkcuda} demonstrate the external-memory and
external-semaphore patterns vkGSplat reuses where the renderer needs
to share images with a Vulkan presentation engine for debug viewers,
even though the production SDG path is headless.

% =====================================================================
\section{Conclusion}
% =====================================================================

This paper began as a software-rendering argument, but the stronger conclusion
is a world-modeling architecture for Physical AI. High-quality data generation
should not be framed as a choice between calibrated simulation and learned
world models. It should be framed as an acquisition--state--reconstruction loop:
physical and sensor processes acquire sparse evidence; explicit 3D/4D state
stores evidence, uncertainty, visibility, and motion; classical, Gaussian, and
neural modules reconstruct or predict; and uncertainty decides when to return to
the oracle. The central principle is that AI should reduce the cost of acquiring,
storing, reconstructing, and querying physical truth, not silently replace it.
The lifecycle matters as much as the loop: SDG can spend real time building a
static world, then amortize that world through many dynamic simulation,
acquisition, and reconstruction episodes. Static Gaussian or radiance caches are
therefore powerful, but only when dynamic lights, material edits, moving
occluders, and sensor effects are represented as explicit invalidation and
refresh events.

vkGSplat is the reference implementation of this principle. It keeps Vulkan as a
portable visual-acquisition contract, treats direct hardware Vulkan acquisition
as a valid source of evidence, and adds a CUDA software acquisition path for
AI-compute co-location, custom evidence, and accelerator portability. If the
implementation reaches its throughput, energy, and downstream accuracy targets,
it will support both the software claim --- that graphics APIs can be decoupled
from graphics hardware when the cost model demands it --- and the broader
Physical AI claim: scalable world modeling should be backend-neutral at
acquisition, compute-first at reconstruction, explicit-state, and oracle-aware.

\appendices
\section{Modernized Amdahl allocation details}
\label{app:amdahl}

This appendix integrates the mathematical core of the modernized Amdahl
argument~\cite{lu2026amdahl}.  Classical Amdahl analysis writes
normalized time as
\begin{equation}
  T_A(N) = (1-P) + \frac{P}{N},
  \label{eq:appendix-amdahl}
\end{equation}
where $P$ is the parallelizable fraction and $N$ is processor
count~\cite{amdahl1967}.  Gustafson's law changes the normalization by
allowing workload scale to grow with the machine~\cite{gustafson1988}.
Both forms are useful, but both are phrased around homogeneous
replication.  vkGSplat needs a different question: how should a fixed
resource budget be allocated between specialized graphics logic and
programmable compute when the workload itself is moving?

Let $x \in [0,1)$ be the resource fraction allocated to specialized
logic, let $R>1$ be the relative efficiency advantage of that specialized
logic on the bounded part of the workload, and let $s \in [0,1]$ be the
value-scalable fraction: the share for which additional effective
compute continues to improve quality, capability, or utility over the
design interval.  The first-order allocation objective is
\begin{equation}
  T(R,s,x) =
  \frac{1-s}{1+(R-1)x} + \frac{s}{1-x}.
  \label{eq:appendix-allocation}
\end{equation}
In plain text, the modernized Amdahl objective is T(R, s, x) =
(1-s)/(1+(R-1)x) + s/(1-x).
The first term is the bounded share helped by specialization; the second
is the scalable share left on programmable compute.  Differentiating,
\begin{equation}
  \frac{\partial T}{\partial x}
  =
  -\frac{(1-s)(R-1)}{\left(1+(R-1)x\right)^2}
  + \frac{s}{(1-x)^2}.
  \label{eq:appendix-derivative}
\end{equation}
The objective is strictly convex on $[0,1)$, so the sign of
$\partial T/\partial x$ at $x=0$
determines whether the optimum is interior or collapses to the boundary:
\begin{equation}
  \left.\frac{\partial T}{\partial x}\right|_{x=0} < 0
  \quad\Longleftrightarrow\quad
  s < 1 - \frac{1}{R}.
  \label{eq:appendix-threshold}
\end{equation}
Thus, for a fixed specialized efficiency ratio $R$, specialization
collapses when
\begin{equation}
  s_c = 1 - \frac{1}{R}.
  \label{eq:appendix-sc}
\end{equation}
Equivalently, for a fixed scalable fraction $s$, the minimum efficiency
ratio needed to justify specialization is
\begin{equation}
  R_c = \frac{1}{1-s}.
  \label{eq:appendix-rc}
\end{equation}
In the interior regime, solving $\partial T/\partial x=0$ yields
\begin{equation}
  x^*
  =
  \frac{\sqrt{\dfrac{(1-s)(R-1)}{s}} - 1}
       {\sqrt{\dfrac{(1-s)(R-1)}{s}} + (R-1)}.
  \label{eq:appendix-optimum}
\end{equation}

\begin{figure}[t]
\centering
\begin{tikzpicture}
\begin{axis}[
  width=\linewidth,
  height=0.58\linewidth,
  xmin=0, xmax=0.97,
  ymin=1, ymax=30,
  domain=0:0.966,
  samples=300,
  xlabel={$s$ (value-scalable fraction)},
  ylabel={$R$ (efficiency ratio)},
  grid=both,
  grid style={draw=black!10},
  major grid style={draw=black!20},
  ticklabel style={font=\scriptsize},
  label style={font=\footnotesize},
  legend style={font=\scriptsize, cells={anchor=west}, draw=black!20},
  legend pos=north west,
]
\addplot[color=plotcyan, very thick] {1/(1-x)};
\addlegendentry{$R_c = 1/(1-s)$}
\draw[dashed, black!70] (axis cs:0.9,1) -- (axis cs:0.9,10);
\draw[dashed, black!70] (axis cs:0,10) -- (axis cs:0.9,10);
\node[anchor=west, font=\scriptsize, align=left] at (axis cs:0.12,21)
  {nonzero\\specialized share};
\node[anchor=west, font=\scriptsize, align=left] at (axis cs:0.53,4.2)
  {collapse\\$x^*=0$};
\node[anchor=east, font=\scriptsize] at (axis cs:0.89,10.8)
  {$s_c=0.9$ for $R=10$};
\end{axis}
\end{tikzpicture}
\caption{Modernized Amdahl threshold.  As the value-scalable fraction
$s$ rises, the relative efficiency ratio required for a nonzero
specialized allocation rises as $R_c=1/(1-s)$.  This is the mathematical
reason vkGSplat treats fixed-function visibility as bounded acquisition
and keeps reconstruction on programmable compute.}
\label{fig:appendix-amdahl-threshold}
\end{figure}

The interpretation for vkGSplat is direct.  Software visibility, shadow,
and low-sample ray kernels may lose constant factors against dedicated
graphics hardware, but those stages are bounded once they have produced
the seed buffers.  Reconstruction is the value-scalable side: temporal
reuse, denoising, super-resolution, neural fields, and model-driven scene
synthesis can keep absorbing additional effective compute.  As $s$ rises,
the threshold in Fig.~\ref{fig:appendix-amdahl-threshold} says that a
fixed-function graphics block must clear an increasingly high efficiency
bar to justify owning a large fraction of the system.

A bandwidth-limited extension strengthens the same conclusion.  In a
Roofline-style view~\cite{williams2009}, a specialized block may fail to
realize its peak efficiency if memory supply is limiting.  Replacing $R$
with an effective ratio
\begin{equation}
  R_{\mathrm{eff}}(x) =
  \frac{R_{\max}}{1 + \gamma R_{\max}x}
  \label{eq:appendix-reff}
\end{equation}
deforms the interior optimum toward smaller $x$, while the first-order
threshold at the origin remains governed by $s_c=1-1/R_{\max}$.  Memory
pressure therefore reinforces the preference for a broad programmable
substrate in the SDG regime.

\section*{Acknowledgments}
The author thanks the maintainers of the open-source projects this
proposal builds on --- the original 3D Gaussian Splatting team, the
Mesa Lavapipe and SwiftShader communities, the Mitsuba\,/\,Dr.Jit
authors, and the Embree and OptiX teams --- whose work makes the
position defended here a credible engineering target rather than a
speculation. The framing of the DLSS argument owes a debt to the body
of NVIDIA technical writing on real-time neural reconstruction; the
errors in interpreting it are the author's alone.

\bibliographystyle{IEEEtran}
\bibliography{references}

\end{document}
